---
title: "Rank-normalized split-Rhat and relative efficiency estimates"
author: "Aki Vehtari, Andrew Gelman, Daniel Simpson, Bob Carpenter, Paul Bürkner"
date: "First version 2017-05-23. Last modified `r format(Sys.Date())`."
encoding: "UTF-8"
output:
  html_document:
    fig_caption: yes
    toc: TRUE
    toc_depth: 2
    number_sections: TRUE
    toc_float:
      smooth_scroll: FALSE
bibliography: rhat_neff.bib
csl: harvard-cite-them-right.csl
---

# Setup {.unnumbered}

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = NA, 
  cache = FALSE,
  fig.width = 10
)
options(width = 78)
```

The code for alternative new *split*-$\widehat{R}$ and $S_{\rm eff}$ versions are in [monitornew.R](monitornew.R) and code for diagnostic plots are in [monitorplot.R](monitorplot.R).

**Load packages**
```{r, comment=NA, message=FALSE, warning=FALSE, results='hide', cache=FALSE}
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
library(bayesplot)
theme_set(bayesplot::theme_default(base_family = "sans"))
library(tidyverse)
library(gridExtra)
library(latex2exp)
source('monitornew.R')
source('monitorplot.R')
```

# Introduction

Split-$\widehat{R}$ and effective sample size $S_{\rm eff}$
(previously $N_{\rm eff}$ or $n_{\rm eff}$), are routinely used
to monitor the convergence of iterative simulations.  $\widehat{R}$
[@Gelman+Rubin:1992; @Brooks+Gelman:1998] and
*split*-$\widehat{R}$ [@BDA3] are based on the ratio of
between and within-chain marginal variances of the simulations. The
effective sample size estimate by @BDA3
is based on marginal variance and autocorrelation estimates from split
chains.

$\widehat{R}$, *split*-$\widehat{R}$, and $S_{\rm eff}$ are well
defined only if the marginal distributions have finite mean and
variance.  and even in that case, they are less stable for
distributions with long tails. To alleviate these problems, we define
*split*-$\widehat{R}$ and $S_{\rm eff}$ using rank normalized values,
empirical cumulative density functions, and small posterior intervals.

[rhat_reff_extra.Rmd](rhat_reff_extra.html) has longer description and
more experiments.

# Review of *split*-$\widehat{R}$ and effective sample size

In this section, we will review split-$\widehat{R}$ and 
effective sample size estimates as implemented in Stan 2.18 [@Stan.2.18].

## *Split*-$\widehat{R}$ {#SplitRhat}

Below, we present the computation of *Split*-$\widehat{R}$ following
@BDA3, but using the notation style of @StanBook. Our recommendation
is to run several chains. $N$ is the number of draws per chain, $M$ is
the number of chains, and $S=MN$ is the total number of draws from all
chains. For each scalar summary of interest $\theta$, we
compute $B$ and $W$, the between- and within-chain variances:
<!--
$$
B = \frac{N}{M-1}\sum_{m=1}^{M}(\overline{\theta}_{.m}-\overline{\theta}_{..})^2,
\;\mbox{ where }\;\;\overline{\theta}_{.m}=\frac{1}{N}\sum_{n=1}^nN \theta_{nm},\;\;
\;\;\overline{\theta}_{..} = \frac{1}{M}\sum_{m=1}^M\overline{\theta}_{.m}\\
W =
 \frac{1}{M}\sum_{m=1}^{M}s_j^2, \;\mbox{ where }\;\; s_m^2=\frac{1}{N-1}
\sum_{n=1}^N (\theta_{nm}-\overline{\theta}_{.m})^2.
$$
-->
$$
B = \frac{N}{M-1}\sum_{m=1}^{M}(\overline{\theta}^{(.m)}-\overline{\theta}^{(..)})^2,
\;\mbox{ where }\;\;\overline{\theta}^{(.m)}=\frac{1}{N}\sum_{n=1}^N \theta^{(nm)},\;\;
\;\;\overline{\theta}^{(..)} = \frac{1}{M}\sum_{m=1}^M\overline{\theta}^{(.m)}\\
W =
 \frac{1}{M}\sum_{m=1}^{M}s_j^2, \;\mbox{ where }\;\; s_m^2=\frac{1}{N-1}
\sum_{n=1}^N (\theta^{(nm)}-\overline{\theta}^{(.m)})^2.
$$

The between-chain variance, $B$, 
also contains the factor $N$ because it is based on the variance of 
the within-chain means, $\overline{\theta}^{(.m)}$, each of which 
is an average of $N$ values $\theta^{(nm)}$.

We can estimate $\mbox{var}(\theta \mid y)$, the marginal posterior variance
of the estimand, by a weighted average of $W$ and $B$, namely
$$
\widehat{\mbox{var}}^+(\theta \mid y)=\frac{N-1}{N}W + \frac{1}{N}B.
$$
This quantity *overestimates* the marginal posterior variance 
assuming the starting distribution is appropriately overdispersed
compared to the target distribution, 
but is *unbiased* under stationarity (that is, if the starting 
distribution equals the target distribution), or in the limit
$N\rightarrow\infty$. To have an overdispersed starting distribution, 
independent Markov chains should be initialized with diffuse starting
values for the parameters.

Meanwhile, for any finite $N$, the within-chain variance
$W$ should *underestimate* $\mbox{var}(\theta \mid y)$ because the 
individual chains haven't had the time to explore all of the target 
distribution and, as a result, will have less variability. 
In the limit as $N\rightarrow\infty$, the expectation of $W$ 
approaches $\mbox{var}(\theta \mid y)$.

We monitor convergence of the iterative simulation to the target
distribution by estimating the factor by which the scale of the 
current distribution for $\theta$ might be reduced if the simulations 
were continued in the limit $N\rightarrow\infty$. This potential 
scale reduction is estimated as
$$
\widehat{R}=
\sqrt{\frac{\widehat{\mbox{var}}^+(\theta \mid y)}{W}},
$$
which declines to 1 as $N\rightarrow\infty$. We call this *split*-$\widehat{R}$
because we are applying it to chains that have been split in half so that $M$ is
twice the number of actual chains. Without splitting, $\widehat{R}$ would get
fooled by non-stationary chains.

We note that *split*-$\widehat{R}$ is also well defined for sequences
that are not Markov-chains. However, for simplicity, we always refer
to 'chains' instead of more generally to 'sequences' as the former is
our primary use case for $\widehat{R}$ measures.

## Effective sample size $S_{\rm eff}$ {#Seff}

If the $N$ simulation draws within each chain were truly
independent, the between-chain variance $B$ would be an unbiased
estimate of the posterior variance, $\mbox{var}(\theta \mid y)$, and we would have
a total of $S = M \times N$ independent simulations from the $M$
chains. In general, however, the simulations of $\theta$ within each
chain will be autocorrelated, and thus $B$ will be larger than 
$\mbox{var}(\theta \mid y)$, in expectation.

A nice introductory reference for analyzing MCMC results in general
and effective sample size in particular is provided by @Geyer:2011
[see also @Geyer:1992].  The particular calculations used by Stan
[@Stan.2.18] follow those for split-$\hat{R}$, which involve both
between-chain (mean) and within-chain calculations
(autocorrelation). They were introduced in the Stan manual
[@StanManual.2.18.0] and explained in more detail in @BDA3.

One way to define effective sample size for correlated simulation
draws is to consider the statistical efficiency of the average of the
simulations $\bar{\theta}^{(..)}$ as an estimate of the posterior mean
$\mbox{E}(\theta \mid y)$. This can be a reasonable baseline even though it is
not the only possible summary and might be inappropriate, for example,
if there is particular interest in accurate representation of
low-probability events in the tails of the distribution.

The effective sample size of a chain is defined in terms of the
autocorrelations within the chain at different lags. The
autocorrelation $\rho_t$ at lag $t \geq 0$ for a chain with joint
probability function $p(\theta)$ with mean $\mu$ and variance
$\sigma^2$ is defined to be
$$
\rho_t 
=
\frac{1}{\sigma^2} \, \int_{\Theta} (\theta^{(n)} - \mu)
(\theta^{(n+t)} - \mu) \, p(\theta) \, d\theta.
$$
This is just the correlation between the two chains offset by $t$
positions.  Because we know $\theta^{(n)}$ and $\theta^{(n+t)}$ have
the same marginal distribution in an MCMC setting, multiplying the
two difference terms and reducing yields
$$
\rho_t
=
\frac{1}{\sigma^2} \, \int_{\Theta} \theta^{(n)} \, \theta^{(n+t)} \, p(\theta) \, d\theta.
$$

The effective sample size of one chain generated by a process with
autocorrelations $\rho_t$ is defined by
$$
N_{\rm eff}
\ = \
\frac{N}{\sum_{t = -\infty}^{\infty} \rho_t}
\ = \
\frac{N}{1 + 2 \sum_{t = 1}^{\infty} \rho_t}.
$$

Effective sample size $N_{\rm eff}$ can be larger than $N$ in case of
antithetic Markov chains, which have negative autocorrelations on odd
lags. Dynamic Hamiltonian algorithm [@Hoffman+Gelman:2014] used in Stan 
can produce $N_{\rm eff}>N$ for parameters which have close to Gaussian 
posterior with low dependency on other parameters (before version 2.18, 
Stan computed $N_{\rm eff}$ so that the maximum reported value was $N$).

\subsection{Estimation of Effective Sample Size}

In practice, the probability function in question cannot be tractably
integrated and thus neither autocorrelation not the effective sample size
can be calculated. Instead, these quantities must be estimated
from the samples themselves. The rest of this section describes an
autocorrelation and split-$\hat{R}$ based effective sample
size estimator, based on multiple split chains. For simplicity, 
each chain will be assumed to be of the same length $N$.

Stan carries out the autocorrelation computations for all lags
simultaneously using Eigen's fast Fourier transform (FFT) package with
appropriate padding; see @Geyer:2011 for more details on using
FFT for autocorrelation calculations.
The autocorrelation estimates $\hat{\rho}_{t,m}$ at lag $t$ from
multiple chains $m \in (1,\ldots,M)$ are combined with the within-chain
variance estimate $W$ and the multi-chain variance estimate
$\widehat{\mbox{var}}^{+}$ introduced in the previous section to
compute the combined autocorrelation at lag $t$ as
$$
\hat{\rho}_t
= 1 - \frac{\displaystyle W - \textstyle \frac{1}{M}\sum_{m=1}^M 
\hat{\rho}_{t,j}}{\widehat{\mbox{var}}^{+}}. \label{rhohat}
$$
If the chains have not converged, the variance estimator
$\widehat{\mbox{var}}^{+}$ will overestimate the true marginal variance 
which leads to an overestimation of the autocorrelation and an 
underestimation of the effective sample size.

Because of the noise in the correlation estimates $\hat{\rho}_t$ as
$t$ increases, typically the truncated sum of $\hat{\rho}_t$ is used.
Negative autocorrelations can happen only on odd lags and by summing
over pairs starting from lag $t=0$, the paired autocorrelation is
guaranteed to be positive, monotone and convex modulo estimator noise
[@Geyer:1992; @Geyer:2011].  Stan 2.18 uses Geyer's initial monotone
sequence criterion. The effective sample size of combined chains is defined as
$$
S_{\rm eff} = \frac{N \, M}{\hat{\tau}},
$$
where
$$
\hat{\tau} = 1 + 2 \sum_{t=1}^{2k+1} \hat{\rho}_t = 
-1 + 2 \sum_{t'=0}^{k} \hat{P}_{t'},
$$
and $\hat{P}_{t'}=\hat{\rho}_{2t'}+\hat{\rho}_{2t'+1}$. Initial
positive sequence estimators is obtained by choosing the largest $k$
such that $\hat{P}_{t'}>0, \quad t' = 1,\ldots,k$. The initial monotone
sequence is obtained by further reducing $\hat{P}_{t'}$ to the minimum
of the preceding ones so that the estimated sequence is monotone.

$S_{\rm eff}$ described here is different from similar formulas in
the literature in that we use multiple chains and between-chain variance in the
computation which typically gives us more conservative claims (lower
values of $S_{\rm eff}$) compared to single chain
estimates, especially when mixing is poor. If the chains are not
mixing at all (e.g. the posterior is multimodal and the chains are
stuck in different modes), then our $S_{\rm eff}$ is close to
the number of chains.

In Stan pre 2.18, a slightly incorrect initial sequence was used, which 
implied that $S_{\rm eff}$ was capped at $S$. As
antithetic behavior $S_{\rm eff} > S$ is not that common or the
effect is small and capping at $S$ can be considered to be
pessimistic, this had negligible effect on any inference.
However, it may have led to an underestimation of Stan's efficiency
when compared to other packages performing MCMC sampling.

# Rank normalized *split*-$\widehat{R}$ and relative efficiency estimates

As *split*-$\widehat{R}$, and $S_{\rm eff}$ are well defined only if
the marginal posteriors have finite mean and variance, we next
describe variants which are well defined for all distributions and
more robust for long tailed distributions.

## Rank normalization

1. Rank: Replace each value $\theta^{(nm)}$ by its rank $r^{(nm)}$. Average
rank for ties are used to conserve the number of unique values of discrete
quantities. Ranks are computed jointly for all draws from all chains.
2. Normalize: Normalize ranks by inverse normal transformation $z^{(nm)} =
\phi^{-1}((r^{(nm)}-1/2)/S)$. 

For continuous variables and $S \rightarrow \infty$, the rank
normalized values are normally distributed and rank normalization
performs non-parametric transformation to normal distribution. Using
normalized ranks instead of ranks directly, has the benefit that the
behavior of $\widehat{R}$ and $S_{\rm eff}$ do not change for normally
distributed $\theta$.

## Rank normalized *Split*-$\widehat{R}$

Rank normalized *Split*-$\widehat{R}$ is computed using the equations
in [Section *Split*-$\widehat{R}$](#SplitRhat) by replacing parameter
values $\theta^{(nm)}$ with rank normalized values $z^{(nm)}$.

## Rank normalized folded *split*-$\widehat{R}$

Both original and rank-normalized *Split*-$\widehat{R}$ can miss if
chains have different scales but the same location. To alleviate this,
we propose to compute rank normalized *folded-split*-$\widehat{R}$
using *folded* split chains by rank normalizing absolute deviations from
median
$$
  {\rm abs}(\theta^{(nm)}-{\rm median}(\theta)).
$$

<!--
Notation for median?
-->

To obtain a single conservative $\widehat{R}$ estimate, we propose to
report the maximum of *rank-normalized-split*-$\widehat{R}$ and
*rank-normalized-folded-split*-$\widehat{R}$ for each parameter.

## Relative efficiency using rank normalized values

In addition of using rank-normalized values for convergence
diagnostics via $\widehat{R}$, we can also compute the corresponding
effective sample size.  This estimate will be well defined even if the
original distribution does not have finite mean and variance. It is
not directly applicable to estimate the Monte Carlo error of the mean
of the original values, but it will provide bijective transformation
invariant estimate of the mixing efficiency of chains. For simplicity
we propose to report relative efficiency values
$$
R_{\rm eff}=\textit{rank-normalized-split-}S_{\rm eff} / S,
$$
where $\textit{split-}S_{\rm eff}$ is computed using equations in
[Section Effective sample size](#Seff) by replacing parameter values
$\theta^{(nm)}$ with rank normalized values $z^{(nm)}$. For $\theta$ which
have close to normal distribution the difference to using the original
values is small. For $\theta$ which has a distribution far from
normal, rank normalization can be seen as near optimal non-parametric
transformation.

The relative efficiency estimate using means and variances of rank
normalized values is a useful measure for relative efficiency of
estimating the bulk (mean and quantiles near median) of the
distribution, and as shorthand term we use term *bulk relative
efficiency* (*bulk*-$R_{\rm eff}$).

We propose to compute the relative efficiency also using *folded*
split chains by rank normalizing absolute deviations from median,
which is useful measure for relative efficiency of estimating the tail
of the distribution. As a shorthand term we use a term *tail relative
efficiency* (*tail*-$R_{\rm eff}$).

## Relative efficiency of cumulative distribution function

Bulk and tail relative efficiency measures introduced above are useful
as overall efficiency measures. Next we introduce relative efficiency
estimate of cumulative distribution function, and later we use that to
introduce relative efficiency of quantiles and local small probability
interval relative efficiency diagnostic.

The quantiles and posterior intervals are often reported quantities,
which are easy to estimate from posterior draws.  Estimating the
relative efficiency of such quantiles thus has high practical
relevance in particular as we observe the relative efficiency for tail
quantiles is often lower than for mean.  $\alpha$-quantile is
$\theta^*$ for which $p(\theta < \theta^*) = \alpha$. This is not in a
form of expectation, and thus $\widehat{R}$ and $S_{\rm eff}$
equations are not directly applicable. Thus we first focus on
relative efficiency of cumulative distribution function $p(\theta <
\theta^*)$ for different values of $\theta^*$. 

For any $\theta^*$ ECDF gives an estimate of cumulative probability
$$
p(\theta<\theta^*) \approx \bar{I} = \frac{1}{S}\sum_{s=1}^S
I(\theta^{(s)}<\theta^*),
$$
where $I()$ is the indicator function. The indicator
function transforms simulation draws to 0's and 1's, and thus the
subsequent computations are bijectively invariant. 

## Relative efficiency of quantiles {#quantile_R_eff}

Assuming the cumulative distribution function $F$ is continuous and smooth near
an $\alpha$-quantile and we know $F$, we could use delta method to
compute a variance estimate for $F_\theta^{-1}(\bar{I})$. Although we
don't usually know $F$, delta method approach reveals that the
variance of $\bar{I}$ is scaled by the (usually unknown) density
$f(\theta^*)$, but the relative efficiency depends only on the
relative efficiency of $\bar{I}$. Thus, we can use relative efficiency
of ECDF (via the indicator function $I(\theta^{(s)}<\theta^*)$) also for the
corresponding quantile estimates.

## Relative efficiency of median and MAD

Since the marginal posterior distributions might not have finite mean
and variance, by default RStan [@RStan.2.17] and RStanARM
[@RStanARM.2.17] report median and median absolute deviation (MAD)
instead of mean and standard error (SE). Median and MAD are well
defined even when the marginal distribution does not have finite mean
and variance.

Median is 50%-quantile, and we can estimate the relative efficiency as
for the other quantiles.  

We can also compute the relative efficiency for the
median absolute deviation (MAD), by computing the relative efficiency
for median of absolute deviations from the median of all draws. The
absolute deviations from the median of all draws are same as
previously defined for folded samples
$$
{\rm abs}(\theta^{(nm)}-{\rm median}(\theta)).
$$
We see that the relative efficiency of MAD is obtained by using the
same approach as for median (and other quantiles) but with the folded
values used also used in
*rank-normalized-folded-split*-$S_{\rm eff}$.

## Monte Carlo error estimates for quantiles

Previously, Stan has reported Monte Carlo standard error estimates for
mean of a quantity. This is valid only if the corresponding
distribution has finite mean and variance. It would be possible to
estimate when mean and variance are finite, but easier option is focus
on median and other quantiles.

Median, MAD and quantiles are well defined even when the distribution
does not have finite mean and variance, and they are asymptotically
normally distributed for distributions which are continuous and
non-zero in the relevant neighborhood. As the delta method of
computing the variance would require knowledge of the normalized
posterior density, we propose to use following approach.

1. Compute quantiles of ${\rm Beta}((R_{\rm eff} Y)+1, R_{\rm eff}(S - Y) + 1)$
distribution, where $Y=\sum_{s=1}^S I(\theta^{(s)}<\theta^*)$. Including
$R_{\rm eff}$ takes into account the relative efficiency of MCMC.
2. Find posterior draws with closest corresponding ranks. For example,
in case of quantiles $\alpha_1$ and $\alpha_2$, find $r_1$ and $r_2$
which are closest to $\alpha_1 S$ and $\alpha_2 S$.
3. Use the corresponding $\theta^{(s)}$ as quantiles from the
error distribution. These quantiles can be used to approximate
Monte Carlo standard error.

## Relative efficiency of small interval probability estimates {#small_interval_R_eff}

We can get more local relative efficiency estimate by considering
small probability intervals. We propose to compute the relative
efficiency for
$$
\bar{I} = p(\hat{Q}_\alpha \le \theta < \hat{Q}_{\alpha+\delta}),
$$
where $\hat{Q}_\alpha$ is an empirical $\alpha$-quantile, $\delta=1/k$
is the length of the interval with some positive integer $k$, and
$\alpha \in (0,\delta,\ldots,1-\delta)$ changes in steps of $\delta$.
Each interval has $S/k$ draws, and the efficiency measures
autocorrelation of an indicator function which is one when the values
are inside the specific interval and zero otherwise. This gives us a
local efficiency measure which does not depend on the shape of the
distribution.

## Rank plots

In addition to using rank-normalized values to compute
*split*-$\widehat{R}$, we rank plots for each chain
instead of trace plots. Rank plots are nothing else than histograms of the 
ranked posterior samples (ranked over all chains) plotted separately
for each chain. If rank plots of all chains look similar, this indicates
good mixing of the chains. As compared to trace plots, they don't tend
to squeeze to a mess in case of long chains.

# Examples

## Cauchy: A distribution with infinite mean and variance

The classic split-Rhat is based on calculating within and between chain
variances. If the marginal distribution of a chain is such that
the variance is not defined (i.e. infinite), the classic split-Rhat is not 
well justified. In this section, we will use the Cauchy distribution 
as an example of such distribution. 

The following Cauchy models are from Michael Betancourt's case study
[Fitting The Cauchy Distribution](https://betanalpha.github.io/assets/case_studies/fitting_the_cauchy.html)

### Nominal parameterization of Cauchy

The nominal Cauchy model with direct parameterization is as follows.

```{r}
writeLines(readLines("cauchy_nom.stan"))
```

#### Default Stan options

Run the nominal model:
```{r fit_nom, cache=TRUE, comment=NA, results='hide'}
fit_nom <- stan(file = 'cauchy_nom.stan', seed = 7878, refresh = 0)
```

Treedepth exceedence and Bayesian Fraction of Missing Information are
dynamic HMC specific diagnostics [@betancourt2017conceptual]. We get
warnings about very large number of transitions after
warmup that exceeded the maximum treedepth, which is likely due to
very long tails of the Cauchy distribution. All chains have low 
estimated Bayesian fraction of missing information also indicating
slow mixing.

```{r}
mon <- monitornew(fit_nom)
print(mon)
which_min_eff <- which.min(mon[1:50, 'Bulk_Reff'])
```

Several Rhat>1.01 and some Rhat<0.1 and the results can not be trusted.
[rhat_reff_extra.Rmd](rhat_reff_extra.html) has results for longer chains.

We can further analyze potential problems using local relative
efficiency and rank plots.  We examine x[`r which_min_eff`], which has
the smallest bulk relative efficiency `r round(min(mon[,'Bulk_Reff']), 2)`.

We examine the relative efficiency in different parts of the
posterior by computing the relative efficiency of small interval
probability estimates (see [Section Relative efficiency of small
interval probability estimates](#small_interval_R_eff)).  Each
interval contains $1/k$ of the draws (e.g., with $k=20$).
The small interval efficiency measures mixing of an indicator
function which indicates when the values are inside the specific
small interval.  This gives us a local efficiency measure which does not
depend on the shape of the distribution.

```{r}
plot_local_reff(fit = fit_nom, par = which_min_eff, nalpha = 20)
```

We see that the efficiency of MCMC is worryingly low in the tails
(which is caused by slow mixing in long tails of Cauchy). Orange ticks
show chains with maximum treedepth.

Alternative way to examine the relative efficiency in different parts
of the posterior, is to compute relative efficiency for quantiles (see
[Section Relative efficiency of quantiles](#quantile_R_eff)). Each
interval has specified proportion of draws, and the efficiency measures
mixing of an indicator function which indicates when the
values are inside the specific interval. 

```{r, cahce=FALSE}
plot_quantile_reff(fit = fit_nom, par = which_min_eff, nalpha = 40)
```

We see that the efficiency of MCMC is worryingly low in the tails
(which is caused by slow mixing in long tails of Cauchy). Orange ticks
show chains with maximum treedepth.

We can further analyze potential problems using rank plots.

```{r}
samp <- as.array(fit_nom)
xmin <- paste0("x[", which_min_eff, "]")
mcmc_hist_r_scale(samp[, , xmin])
```

The chains clearly have different rank plots.


### Alternative parameterization of Cauchy

Next we examine alternative parameterization and consider Cauchy as a
scale mixture of Gaussian distributions. The model has two parameters
and the Cauchy distributed x's can be computed from those. In addition
of improved sampling performance, the example illustrates that focus
of diagnostics matter.

```{r}
writeLines(readLines("cauchy_alt_1.stan"))
```

Run the alternative model:

```{r fit_alt1, cache=TRUE, comment=NA, results='hide'}
fit_alt1 <- stan(file = 'cauchy_alt_1.stan', seed = 7878, refresh = 0)
```

There are no warnings, and the sampling is much faster.

```{r}
mon <- monitornew(fit_alt1)
print(mon)
which_min_eff <- which.min(mon[101:150, 'Bulk_Reff'])
```

All Rhat<1.01 and Rhat>0.1 and the MCMC works much better with the alternative parameterization.
[rhat_reff_extra.Rmd](rhat_reff_extra.html) has results for other alternative parameterizations, too.

We can further analyze potential problems using local relative efficiency and
rank plots.  We examine x[`r which_min_eff`], which has the smallest bulk
relative efficiency `r round(mon[which_min_eff, 'Bulk_Reff'], 2)`.

We examine the relative efficiency in different parts of the
posterior by computing the relative efficiency of small interval
probability estimates.

```{r}
plot_local_reff(fit = fit_alt1, par = which_min_eff + 100, nalpha = 20)
```

The relative efficiency is good in all parts of the posterior.

We examine also the relative efficiency of different quantile estimates.

```{r}
plot_quantile_reff(fit = fit_alt1, par = which_min_eff + 100, nalpha = 40)
```

```{r}
samp <- as.array(fit_alt1)
xmin <- paste0("x[", which_min_eff, "]")
mcmc_hist_r_scale(samp[, , xmin])
```

Looks better than for the nominal parameterization. There are still
some differences in tails, which was also reflected in tail quantity
relative efficiencies.


### Half-Cauchy with nominal parameterization

Half-Cauchy priors are common and, for example, in Stan usually set
using the nominal parameterization. However, when the constraint 
`<lower=0>` is used, Stan does the sampling automatically
in the unconstrained `log(x)` space, which changes the geometry
crucially.

```{r}
writeLines(readLines("half_cauchy_nom.stan"))
```

Run the half-Cauchy with nominal parameterization (and positive constraint).

```{r fit_half_nom, cache=TRUE, comment=NA, results='hide'}
fit_half_nom <- stan(file = 'half_cauchy_nom.stan', seed = 7878, refresh = 0)
```

There are no warnings, and the sampling is much faster than for the
Cauchy nominal model.

```{r}
mon <- monitornew(fit_half_nom)
print(mon)
```

All Rhat<1.01 and Rhat>0.1 and the MCMC works much better than the
nominal parameterization in the unconstrained case. We see that the
Stan's automatic transformation of constraint parameters can have a
big effect in the sampling performance.

[rhat_reff_extra.Rmd](rhat_reff_extra.html) has more diagnostic plots
and alternative parameterization.


## Hierarchical model: Eight Schools

Eight Schools data is classic example for hierarchical models (Section
5.5, @BDA3), which despite the apparent simplicity nicely illustrates
the typical problems in inference for hierarchical models.  The Stan
models below are from Michael Betancourt's case study on [Diagnosing
Biased Inference with
Divergences](http://mc-stan.org/users/documentation/case-studies/divergences_and_bias.html).

### A Centered Eight Schools model

```{r, comment=NA}
writeLines(readLines("eight_schools_cp.stan"))
```

#### Centered parameterization default MCMC options

Run centered parameterization model with default MCMC options.
```{r fit_cp, cache=TRUE, comment=NA, results='hide'}
input_data <- read_rdump("eight_schools.data.R")
fit_cp <- stan(
  file = 'eight_schools_cp.stan', data = input_data,
  iter = 2000, chains = 4, seed=483892929, refresh = 0
)
```

We do observe divergent transitions, which is sensitive diagnostic for
this model and we observer also low BFMI, so we already know that the
results can't be trusted.  We can use Rhat and Reff diagnostics also
for other MCMC algorithms, but they can also be helpful to recognize
problematic parts of the posterior.

```{r}
mon <- monitornew(fit_cp)
print(mon)
```

Several Rhat>1.01 and tau Rhat<0.1 and the results can not be trusted.
[rhat_reff_extra.Rmd](rhat_reff_extra.html) has results for longer chains.

We examine the relative efficiency in different parts of the posterior
by computing the relative efficiency of small interval probability
estimates for `tau`. We can plot showing either quantiles or parameter
values at the vertical axis.

```{r}
plot_local_reff(fit = fit_cp, par = "tau", nalpha = 20)
```

```{r}
plot_local_reff(fit = fit_cp, par = "tau", nalpha = 20, rank = FALSE)
```

We see that MCMC has difficulties in exploring small `tau` values. As
the efficiency for estimating small `tau` values is practically zero,
we may assume that we may miss substantial amount of posterior mass
and get biased estimates. Red ticks which show chains with divergences
have concentrated to small `tau` values, indicating problems exploring
even smaller values which is likely to cause bias.

We examine also the relative efficiency of different quantile
estimates. We can plot showing either quantiles or parameter values at
the vertical axis.

```{r}
plot_quantile_reff(fit = fit_cp, par = 2, nalpha = 40)
```

```{r}
plot_quantile_reff(fit = fit_cp, par = 2, nalpha = 40, rank = FALSE)
```

Most of the quantile estimates have worryingly low relative efficiency.

Rank plot visualisation of `tau`.

```{r}
samp_cp <- as.array(fit_cp)
mcmc_hist_r_scale(samp_cp[, , "tau"])
```

We observe clear problems.


### Non-centered Eight Schools model

For hierarchical models, non-centered parameterization often works better
```{r, comment=NA}
writeLines(readLines("eight_schools_ncp.stan"))
```

#### Non-centered parameterization default MCMC options plus `adapt_delta=0.95`

Run non-centered parameterization model with default options plus `adapt_delta=0.95`

```{r fit_ncp2, cache=TRUE, comment=NA, results='hide'}
fit_ncp2 <- stan(
  file = 'eight_schools_ncp.stan', data = input_data,
  iter = 2000, chains = 4, control = list(adapt_delta = 0.95), 
  seed = 483892929, refresh = 0
)
```

We get zero divergences with `adapt_delta = 0.95`.

```{r}
mon <- monitornew(fit_ncp2)
print(mon)
```

All Rhat<1.01 and Rhat>0.1 and the MCMC works much better with the
non-centered parameterization.

We examine the relative efficiency in different parts of the
posterior by computing the relative efficiency of small interval
probability estimates for `tau`.
```{r}
plot_local_reff(fit = fit_ncp2, par = 2, nalpha = 20)
```

Small `tau` values are still more difficult to explore, but the relative
efficiency is in a good range.

Check with a finer resolution
```{r}
plot_local_reff(fit = fit_ncp2, par = 2, nalpha = 50)
```

Also with a finer resolution the efficiency of MCMC is good for
different values of tau.

We examine also the relative efficiency of different quantile estimates.
```{r}
plot_quantile_reff(fit = fit_ncp2, par = 2, nalpha = 40)
```

Rank plot visualisation of `tau`, which has the smallest relative
efficiency 0.62 among parameters.
```{r}
samp_ncp2 <- as.array(fit_ncp2)
mcmc_hist_r_scale(samp_ncp2[, , 2])
```

Seems ok.

# References {.unnumbered}

<div id="refs"></div>

# Original Computing Environment {.unnumbered}

```{r, comment=NA}
makevars <- file.path(Sys.getenv("HOME"), ".R/Makevars")
if (file.exists(makevars)) {
  writeLines(readLines(makevars)) 
}
```

```{r, comment=NA}
devtools::session_info("rstan")
```

