---
title: "Rank-normalized split-Rhat and relative efficiency estimates"
author: "Aki Vehtari, Andrew Gelman, Daniel Simpson, Bob Carpenter, Paul Bürkner"
date: "August 2018"
encoding: "UTF-8"
output:
  html_document:
    fig_caption: yes
    toc: TRUE
    toc_depth: 2
    number_sections: TRUE
    toc_float:
      smooth_scroll: FALSE
bibliography: rhat_neff.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment=NA, 
  cache=TRUE,
  fig.width = 10
)
```

# Introduction

Split-$\widehat{R}$ and effective sample size, are routinely used
monitor the convergence of iterative simulations.  $\widehat{R}$
[@Gelman+Rubin:1992; @Brooks+Gelman:1998] and
*split*-$\widehat{R}$ [@BDA3] are based on the ratio of
between and within-sequence marginal variances of the simulations. The
effective sample size estimate ($N_{\rm eff}$) by @BDA3
is based on marginal variance and autocorrelation estimates from split
chains.

$\widehat{R}$, *split*-$\widehat{R}$, and $N_{\rm eff}$ are well
defined only if the marginal posteriors have finite mean and variance,
and even in that case they are less stable for distributions with long
tails.

To alleviate these problems, we propose

 - to compute *split*-$\widehat{R}$ and $N_{\rm eff}$ using rank normalized values,
quantiles and posterior intervals.

We also propose

 - to diagnose convergence and relative efficiency of scale of the marginal
distribution in addition to the usual location estimate (mean or
median), which will improve reliability of Monte Carlo error
estimates,
 - to use relative efficiency of several quantile or small interval probability estimates as local relative efficiency measure which can reveal local sampling problems,
 - to use rank plots as a more compact alternative to trace plots.

# Review of *split*-$\widehat{R}$ and effective sample size

We review split-$\widehat{R}$ and effective sample size estimated as
implemented in Stan 2.18 [@Stan:JSS:2017, @Stan.2.18].

## *Split*-$\widehat{R}$

Here are the formulas, with the presentation taken from @BDA3.  For
each scalar summary of interest $\psi$, we compute $B$ and $W$, the
between- and within-sequence variances:
$$
B = \frac{n}{m-1}\sum_{j=1}^{m}(\overline{\psi}_{.j}-\overline{\psi}_{..})^2,
\;\mbox{ where }\;\;\overline{\psi}_{.j}=\frac{1}{n}\sum_{i=1}^n \psi_{ij},\;\;
\;\;\overline{\psi}_{..} = \frac{1}{m}\sum_{j=1}^m\overline{\psi}_{.j}\\
W =
 \frac{1}{m}\sum_{j=1}^{m}s_j^2, \;\mbox{ where }\;\; s_j^2=\frac{1}{n-1}
\sum_{i=1}^n (\psi_{ij}-\overline{\psi}_{.j})^2.
$$

The between-sequence variance, $B$, contains a factor of $n$ because
it is based on the variance of the within-sequence means,
$\overline{\psi}_{.j}$, each of which is an average of $n$ values
$\psi_{ij}$.

We can estimate $\mbox{var}(\psi|y)$, the marginal posterior variance
of the estimand, by a weighted average of $W$ and $B$, namely
$$
\widehat{\mbox{var}}^+(\psi|y)=\frac{n-1}{n}W + \frac{1}{n}B.
$$
This quantity *overestimates*
the marginal posterior variance assuming the starting distribution
is appropriately overdispersed, but is *unbiased* under
stationarity (that is, if the starting distribution equals the target
distribution), or
in the limit $n\rightarrow\infty$.
To have overdispersed starting distribution, independent Markov chains
should be initialized with diffuse starting values for the parameters.

Meanwhile, for any finite $n$, the ``within'' variance
$W$ should be an *underestimate*
of $\mbox{var}(\psi|y)$ because the individual sequences have not had
time to range over all of the target distribution and, as a result,
will have less variability; in the limit as $n\rightarrow\infty$, the
expectation of $W$ approaches $\mbox{var}(\psi|y)$.


We monitor convergence of the iterative simulation by estimating
the factor by which the scale of the current distribution for $\psi$
might be reduced if the simulations were continued in the limit
$n\rightarrow\infty$.  This potential scale reduction is estimated for
monitoring convergence of iterative simulation by
$$
\widehat{R}=
\sqrt{\frac{\widehat{\mbox{var}}^+(\psi|y)}{W}},
$$
which declines to 1 as $n\rightarrow\infty$.

We call this *split*-$\widehat{R}$ because we are applying it to
chains that have been split in half.  Without splitting, $\widehat{R}$
would get fooled by nonstationary chains.

## Effective sample size

If the $n$ simulation draws within each sequence were truly
independent, then the between-sequence variance $B$ would be an unbiased
estimate of the posterior variance, $\mbox{var}(\psi|y)$, and we would have
a total of $mn$ independent simulations from the $m$
sequences.  In general, however, the simulations of $\psi$ within each
sequence will be autocorrelated, and $B$ will be larger than $\mbox{var}(\psi|y)$,
in expectation.

A nice introductory reference for analyzing MCMC results in general
and effective sample size in particular is [@Geyer:2011; see also @Geyer:1992].  The
particular calculations used by Stan [@Stan.2.18] follow those for split-$\hat{R}$,
which involve both cross-chain (mean) and within-chain calculations
(autocorrelation); they were introduced in Stan manual and explained
in more detail in @BDA3.

One way to define effective sample size for correlated simulation
draws is to consider the statistical efficiency of the average of the
simulations $\bar{\psi}_{..}$, as an estimate of the posterior mean,
$\mbox{E}(\psi|y)$.  This can be a reasonable baseline even though is
not the only possible summary and might be inappropriate, for example,
if there is particular interest in accurate representation of
low-probability events in the tails of the distribution.

The effective sample size of a sequence is defined in terms of the
autocorrelations within the sequence at different lags.  The
autocorrelation $\rho_t$ at lag $t \geq 0$ for a chain with joint
probability function $p(\theta)$ with mean $\mu$ and variance
$\sigma^2$ is defined to be
$$
\rho_t
=
\frac{1}{\sigma^2} \, \int_{\Theta} (\theta^{(n)} - \mu)
(\theta^{(n+t)} - \mu) \, p(\theta) \, d\theta.
$$
This is just the correlation between the two chains offset by $t$
positions.  Because we know $\theta^{(n)}$ and $\theta^{(n+t)}$ have
the same marginal distribution in an MCMC setting, multiplying the
two difference terms and reducing yields
$$
\rho_t
=
\frac{1}{\sigma^2} \, \int_{\Theta} \theta^{(n)} \, \theta^{(n+t)} \, p(\theta) \, d\theta.
$$

The effective sample size of $N$ samples generated by a process with
autocorrelations $\rho_t$ is defined by
$$
N_{\rm eff}
\ = \
\frac{N}{\sum_{t = -\infty}^{\infty} \rho_t}
\ = \
\frac{N}{1 + 2 \sum_{t = 1}^{\infty} \rho_t}.
$$

Effective sample size $N_{\rm eff}$ can be larger than $N$ in case of
antithetic Markov chains, which have negative autocorrelations on odd
lags. Dynamic Hamiltonian algorithm [@Hoffman+Gelman:2014] used in Stan can produce $N_{\rm
eff}>N$ for parameters which have close to Gaussian posterior with low
dependency on other parameters (Stan before 2.18 computed $N_{\rm
eff}$ so that the maximum reported value was $N$).

\subsection{Estimation of Effective Sample Size}

In practice, the probability function in question cannot be tractably
integrated and thus the autocorrelation cannot be calculated, nor the
effective sample size.  Instead, these quantities must be estimated
from the samples themselves.  The rest of this section describes a
autocorrelations and split-$\hat{R}$ based effective sample
size estimator, based on multiple split chains. For simplicity, each chain
$\theta_m$ will be assumed to be of length $N$.

Stan carries out the autocorrelation computations for all lags
simultaneously using Eigen's fast Fourier transform (FFT) package with
appropriate padding; see @Geyer:2011 for more detail on using
FFT for autocorrelation calculations.
The autocorrelation estimates $\hat{\rho}_{t,m}$ at lag $t$ from
multiple chains $m \in (1,\ldots,M)$ are combined with within-sample
variance estimate $W$ and multi-chain variance estimate
$\widehat{\mbox{var}}^{+}$ introduced in the previous section to
compute the combined autocorrelation at lag $t$ as
$$
\hat{\rho}_t
= 1 - \frac{\displaystyle W - \textstyle \frac{1}{M}\sum_{m=1}^M \hat{\rho}_{t,m}}{\widehat{\mbox{var}}^{+}}. \label{rhohat}
$$
If the chains have not converged, the variance estimator
$\widehat{\mbox{var}}^{+}$ will overestimate variance, leading to an
overestimate of autocorrelation and an underestimate effective sample
size.

Because of the noise in the correlation estimates $\hat{\rho}_t$ as
$t$ increases, typical truncated sum of $\hat{\rho}_t$ is used.
Negative autocorrelations can happen only on odd lags and by summing
over pairs starting from lag 0, the paired autocorrelation is
guaranteed to be positive, monotone and convex modulo estimator noise
[@Geyer:1992; @Geyer:2011].  Stan uses Geyer's initial monotone
sequence criterion. The effective sample size estimator is defined as
$$
\hat{N}_{\rm eff}
=
\frac{MN}
     {\hat{\tau}},
$$
where
$$
\hat{\tau} = 1 + 2 \sum_{t=1}^{2m+1} \hat{\rho}_t = -1 + 2 \sum_{t'=0}^{m} \hat{P}_{t'},
$$
where $\hat{P}_{t'}=\hat{\rho}_{2t'}+\hat{\rho}_{2t'+1}$. Initial
positive sequence estimators is obtained by choosing the largest $m$
such that $\hat{P}_{t'}>0, \quad t' = 1,\ldots,m$. The initial monotone
sequence is obtained by further reducing $\hat{P}_{t'}$ to the minimum
of the preceding ones so that the estimated sequence is monotone.

Our $\hat{N}_{\rm eff}$ is slightly different from similar formulas in
the literature in that we use also between-sequence variance in the
computation which typically gives us more conservative claims (lower
values of $N_{\rm eff}$) compared to purely between-sequence
estimates, especially when mixing is poor. If the chains are not
mixing at all (e.g. the posterior is multimodal and the chains are
stuck in different modes), then our $\hat{N}_{\rm eff}$ is close to
the number of chains.

In Stan pre 2.18 slightly wrong initial sequence was used, which had
the effect that $\hat{N}_{\rm eff}$ estimate was capped at $N$. As
antithetic behavior $\hat{N}_{\rm eff}>N$ is not that common or the
effect is small and capping at $N$ can be considered to be
pessismistic, this had negligible effect on any inference.

# New *split*-$\widehat{R}$ and relative efficiency estimate

As *split*-$\widehat{R}$, and $N_{\rm eff}$ are well defined only if
the marginal posteriors have finite mean and variance, we next
describe approaches which are well defined for all distributions and
more robust for long tailed distributions.

## Rank normalization

<!--
*Split*-$\widehat{R}$ assumes that the quantities being considered
have finite mean and variance and it's more stable for distributions
close to Gaussian.  Target distribution can be far from normal and for
example Cauchy distribution has non-fimite mean and  variance,
and thus *split*-$\widehat{R}$ is not well defined. To alleviate
those we propose to compute *split*-$\widehat{R}$ using rank
normalized values.
-->

Rank normalization is commonly used in Spearman's rank correlation
coefficient, but also to transform predictor values in epidemiology.
Rank normalization replaces each draw by its rank (or fractional rank
which the rank divided by the total number of observations). We use
average rank for ties to conserve the number of values the same in
case of discrete quantities. Rank normalization is made jointly for
all draws from all chains.

Rank normalization makes the subsequent computations well defined and
invariant under bijective transformations. This means that we get same
results, for example, if we use unconstrained or constrained
parameters. Furthermore, to improve sensitivity to scale differences
between chains, we consider folded rank-normalization which uses rank
distance from the median.

\paragraph{Folded split}

*Split*-$\widehat{R}$ can miss if chains have different scales but
have the same location. To alleviate this we propose to to compute
*Split*-$\widehat{R}$ also for folded rank normalized values.
Folded values are obtained by absolute deviation transformation
\begin{equation}
  {\rm abs}(y-{\rm median}(y)),
\end{equation}
where median is compute from all draws from all chains. Folded rank
normalized values are then obtained by rank normalizing the folded
values.

## *Combined split*-$\widehat{R}$

For simplicity, we propose to report the maximum of
*rank-normalized-split*-$\widehat{R}$ and
*folded-rank-normalize-split*-$\widehat{R}$ values.

## Relative efficiency of mean rank

In additional of using rank-normalized values for convergence
diagnostics, we can compute also effective sample size using them.
This is estimate will be well defined even if the original
distribution does not have finite mean and variance, but it is not
directly applicabe to estimate the Monte Carlo error of the
expectation of the original value. We could transform quantiles the
Monte Carlo error of the mean rank normalized value back to the
original scale, but for simplicity we propose to report (bijective)
transformation invariant relative efficiency values
$$
R_{\rm eff}=\textit{split-rank-normalized-}N_{\rm eff} / N$,
where $N$ is the total number of draws.

## Relative efficiency of median and MAD

Since the marginal posterior distributions might not have finite mean
and variance, by default RStan [@RStan.2.17] and RStanARM
[@RStanARM.2.17] report median and median absolute deviation (MAD)
instead of mean and standard error (SE). Median and MAD are well
defined also when the marginal distribution does not have finite mean
and variance.

If the median of $x$ is $\tilde{x}$, then by definition $p(x < \tilde{x}) = 0.5$. 
We can't present this as a simple expectation, but we can estimate it by
$\frac{1}{S}\sum I_{x<x'}$. The indicator function transforms simulation 
draws to 0's and 1's, and thus the subsequent computations are
bijectively invariant. Instead of estimating the efficiency of the median
directly, we can estimate the efficiency of $p(x < \hat{Q}_{0.5})$,
where $\hat{Q}_{0.5}$ is the empirical median from the combined
draws. This should be good approximation, if the efficiency for 
$p(x < \hat{Q}_{\alpha})$ is smoothly changing around $\alpha = 0.5$.

Compared to relative efficiency of mean rank, this is more local as it
measures only how the simulation jumps around median.

We can also compute the approximative relative efficiency for the
median absolute deviation (MAD), by computing the relative efficiency
for median of absolute deviations from the median of all draws.  The
absolute deviations from the median of all draws are same as
previously defined folded samples
$$
  {\rm abs}(y-{\rm median}(y)).
$$
We see that relative efficiency of MAD is obtained by using the same
approach as for median but using the folded values used also in
$folded-split-rank-normalized*-$\widehat{R}$.

Previously Stan has reported $N_{\rm eff}$ only for the posterior
location (previously mean), but we demonstrate that it is good to
report the relative efficiency also for scale (we propose to use MAD).

## Relative efficiency of quantiles

We can compute the relative efficiency also for other quantiles than
the median. Some tail quantiles such as the 5\% quantile or the 95\%
quantile are often used to construct credible intervals. Estimating
the relative efficiency of such quantiles thus has high practical
relevance in particular as we observe the relative efficiency for tail
quantiles is often lower than for mean or median. By computing
relative efficiency for many quantiles we can obtain local efficiency
measure demonstrated in the experiments.

## Relative efficiency of small interval probability estimates

We can get even more local relative efficiency estimate by considering small intervals.
We propose to compute the relative efficiency for $p(\hat{Q}_\alpha
\le x < \hat{Q}_{\alpha+\delta})$, where $\hat{Q}_\alpha$ is the empirical
$\alpha$-quantile, $\delta=1/M$ is the length of the interval with some
positive integer $M$ chosen by the researcher, and $\alpha \in (0,\delta,\ldots,1-\delta)$ changes in steps of $\delta$. 
Each interval has $N/M$ draws, and the efficiency measures 
autocorrelation of an indicator function which
indicates when the values are inside the specific interval. This gives
us a local efficiency measure which does not depend on the shape of the
distribution. We demonstrate its usefulness as a diagnostic and
educational tool to teach why expectations of different
functionals may have different $N_{\rm eff}$.

## Local efficiency for distributional approximations

We note that we can plot similar local relative efficiency plots also
for distributional approximations (like Laplace and ADVI
[@Kucukelbir+etal:ADVI:2017] approximations), using Pareto smoothed
importance sampling based effective sample size estimates
[@Vehtari+etal:PSIS:2017]. There will be another case study for this.

## Rank plots

In additional of using rank-normalized values to compute
*split*-$\widehat{R}$, we propose to use rank plots for each chain
instead of traces. These can be used instead of trace plots, which in
case of long chains tend to squeeze to a mess. We illustrate that rank
plots can be used as a more compact alternative to trace plots.

## Finite mean and variance of a marginal

We intentionally separate MCMC diagnostic based on generic and stable
rank-normalized relative efficiency $\hat{n}_{\rm eff}/n$ and Monte
Carlo error estimates for the desired expectations, which is
applicable only when mean and variance exist. We could use Pareto $k$
diagnostic to estimate the existance of mean and variance
[@Vehtari+etal:PSIS:2017]. There will be a separate case study for
diagnostics to check when certain marginal has finite mean and
variance needed in order to report posterior mean and MCSE.

## Abbreviations

In the code and figures the following abbreviations have been used

 - Rhat = classic no-split-Rhat
 - sRhat = classic split-Rhat
 - zsRhat = rank-normalized split-Rhat
   + all draws are jointly ranked, z-transformed and put back to the chains
   + can detect differences in location and trends
 - fzsRhat = folded rank-normalized split-Rhat
   + all draws are jointly "folded" by computing absolute deviation
     from median, ranked, z-transformed and put back to the chains
   + can detect differences in scales
 - neff = no-split effective sample size
 - reff = neff / N
 - zsneff = rank-normalized split effective sample size
   + estimates the efficiency of mean estimate for rank normalized values
 - zsreff = zsneff / N
 - fzsneff = folded rank-normalized split effective sample size
   + estimates the efficiency of rank normalized *mean* absolute deviation
 - fzsreff = fzsneff / N
 - medsneff = median split effective sample size
   + estimates the efficiency of median
   + approximated using efficiency for $p(x < \hat{Q}_{0.5})$
 - medsreff = medsneff / N
 - madsneff = mad split effective sample size
   + estimates the efficiency of median absolute deviation
 - madsreff = madsneff / N

## Proposed change in Stan

The proposal is to switch in Stan

 - from split-$\widehat{R}$ (sRhat) to the maximum of 
 rank-normalized-split-$\widehat{R}$ and 
 folded-rank-normalized-split-$\widehat{R}$ that is max(zsRhat, fzsRhat)
 - from classic effective sample size estimate (neff), 
 which currently doesn't use chain splitting, to 
 rank-normalized-split-$R_{\rm eff}$ (zsreff) and 
 folded-rank-normalized-split-$R_{\rm eff}$ (fzsreff)
 - if computing median and MAD_SD, report corresponding 
 $R_{\rm eff}$'s (medsreff and madsreff)

Justifications for the change are

- Rank normalization makes zsRhat and zsreff well defined for all
  distributions, bijective transformation invariant and more stable
  than sRhat and neff. Adding folded version fzsRhat helps to detect
  differences in scale.

- CDF based medsreff and masreff are well defined for all
  distributions, bijective transformation invariant, more stable than
  neff, and focuses on the median and mad_sd we are reporting.

In displays we propose eventually to use Rhat to denote the new
version, but to make it more clear that rank-normalized efficiency is
different we would switch to use R_eff and display relative
efficiency. The relative efficiency is also easier to check for
low values as we don't need to compare it to the total number of draws.

## Warning thresholds

Based on the experiments, more strict convergence diagnostic and
relative efficiency warning limits could be used.

We propose the following warning thresholds, although additional
experiments would be useful:

 - 1.01 or 1.02 for new Rhat max(zsRhat, fzsRhat)
 - 0.1 for new relative efficiency (zsreff, fzsreff, medsreff, madsreff)

Plots shown in the upcoming section have dashed lines based on these 
thresholds (in continuous plots, a dashed line at 1.005 is plotted instead 
of 1.01, as values larger than that are usually rounded in our summaries to 1.01).

# Experiments

We demonstrated the proposed approached with several simulation studies.

## Setup

**Load packages**
```{r, comment=NA, message=FALSE, warning=FALSE, results='hide'}
library(rstan)
library(bayesplot)
library(tidyverse)
library(gridExtra)
source('monitornew.R')
rstan_options(auto_write = TRUE)
```

```{r monitor_simple}
monitor_simple <- function(sims, ...) {
  out <- monitornew(sims, warmup = 0, probs = 0.5, print = FALSE, ...)
  out <- as.data.frame(out)
  out$par <- seq_len(nrow(out))
  out
}
```

## Independent draws from normal distribution with additional trend, shift or scaling 

This part focuses on diagnostics for

 - all chains have a trend and similar marginal distribution
 - one of the chains has different mean
 - one of the chains has lower marginal variance
 
To simplify, in this part independent draws are used as a proxy for
very efficient MCMC and we sample draws from a standard-normal
distribution. We will discuss the behavior for non-Gaussian
distributions later.

### Trend: All chains from the same $N(0,1)$ distribution plus linear trend added to all chains.

```{r}
conds <- expand.grid(
  iters = c(250, 1000, 4000), 
  trend = c(0, 0.25, 0.5, 0.75, 1),
  rep = 1:10
)
res <- vector("list", nrow(conds))
chains = 4
for (i in 1:nrow(conds)) {
  iters <- conds[i, "iters"]
  trend <- conds[i, "trend"]
  rep <- conds[i, "rep"]
  r <- array(rnorm(iters * chains), c(iters, chains))
  r <- r + seq(-trend, trend, length.out = iters)
  rs <- monitor_simple(r)
  res[[i]] <- cbind(iters, trend, rep, rs)
}
res <- bind_rows(res)
```

If we don't split chains, Rhat misses the trends if all chains still have
similar marginal distribution.
```{r}
ggplot(data=res, aes(y=Rhat, x=trend)) + 
  geom_point() + 
  geom_jitter() + 
  facet_grid(. ~ iters) + 
  geom_hline(yintercept = 1.005, linetype = 'dashed') + 
  geom_hline(yintercept = 1) + 
  ggtitle('Rhat without splitting chains')
```

Split-Rhat can detect trends, even if the marginals of the chains are similar.
```{r}
ggplot(data=res, aes(y=zsRhat, x=trend)) + 
  geom_point() + geom_jitter() + 
  facet_grid(. ~ iters) + 
  geom_hline(yintercept = 1.005, linetype = 'dashed') + 
  geom_hline(yintercept = 1) + 
  ggtitle('Split-Rhat')
```

**Result:** Split-Rhat is useful for detecting non-stationarity in
 the chains. If we use a threshold of $1.01$, we can detect trends which have
 acount for 2% of the total marginal variance. If we use a threshold of $1.1$, we
 can detect trends which acount for 30% of the total marginal variance.

Relative efficiency (effective sample size divided by the number of
draws) is based on split Rhat and within-chain autocorrelation.

```{r}
ggplot(data=res, aes(y=zsreff, x=trend)) + 
  geom_point() + 
  geom_jitter() + 
  facet_grid(. ~ iters) + 
  geom_hline(yintercept = c(0,1)) + 
  geom_hline(yintercept = 0.1, linetype = 'dashed') + 
  ggtitle('Relative efficiency (zsreff)') + 
  scale_y_continuous(breaks = seq(0,1,by=0.25))
```

**Result:** Split-Rhat is more sensitive to trends in small sample
 sizes, but relative efficiency is more sensitive with larger samples
 sizes (as autocorrelations can be estimated more accurately).

**Advice:** If in doubt, run longer chains for more accurate estimates.

### Shifted: Three chains from $N(0,1)$ and one chain from $N(a,1)$

Next we investigate the sensitivity to detect if one of the chains has
not converged to the same distribution as the others, but has a different mean.

```{r}
conds <- expand.grid(
  iters = c(250, 1000, 4000), 
  shift = c(0, 0.25, 0.5, 0.75, 1),
  rep = 1:10
)
res <- vector("list", nrow(conds))
chains = 4
for (i in 1:nrow(conds)) {
  iters <- conds[i, "iters"]
  shift <- conds[i, "shift"]
  rep <- conds[i, "rep"]
  r <- array(rnorm(iters * chains), c(iters, chains))
  r[, 1] <- r[, 1] + shift
  rs <- monitor_simple(r)
  res[[i]] <- cbind(iters, shift, rep, rs)
}
res <- bind_rows(res)
```

```{r}
ggplot(data=res, aes(y=zsRhat, x=shift)) + 
  geom_point() + 
  geom_jitter() + 
  facet_grid(. ~ iters) + 
  geom_hline(yintercept = 1.005, linetype = 'dashed') + 
  geom_hline(yintercept = 1) + 
  ggtitle('Split-Rhat')
```

**Result:** If we use a threshold of $1.01$, we can detect shift with
  a magnitude of one third of the marginal standard deviation. If we use
  a threshold of $1.1$, we can detect shift with a magnitude equal to
  the marginal standard deviation. The rank plot can be used to visualize
  where the problem is.

```{r}
ggplot(data=res, aes(y=zsreff, x=shift)) + 
  geom_point() +
  geom_jitter() + 
  facet_grid(. ~ iters) + 
  geom_hline(yintercept = c(0,1)) + 
  geom_hline(yintercept = 0.1, linetype = 'dashed') + 
  ggtitle('Relative efficiency (zsreff)') + 
  scale_y_continuous(breaks = seq(0,1,by=0.25))
```

**Result:** The relative efficiency is not as sensitive, but a shift
with a magnitude of half the marginal standard deviation will lead to
very low efficiency when sample size increases.

Rank plot visualisation for a case with 4 chains and 250 draws per
chain, and shift = 0.5.
```{r}
iters = 250
chains = 4
shift = 0.5
r <- array(rnorm(iters * chains), c(iters, chains))
r[,1] <- r[,1] + shift
colnames(r) <- 1:4
```

```{r mcmc_hist_r_scale}
mcmc_hist_r_scale <- function(x, nbreaks = 50) {
  max <- prod(dim(x)[1:2])
  mcmc_hist(
    r_scale(x), 
    breaks = seq(0, max, by = max / nbreaks) + 0.5
  )
}
```

```{r}
mcmc_hist_r_scale(r)
```

Rhat is less than $1.05$, but the rank plot clearly shows the location of the problem (first chain).

### Scaled: Three chains from $N(0,1)$ and one chain from $N(0,a)$

Next we investigate the sensitivity to detect if one of the chains has
not converged to the same distribution as the others but has lower
marginal variance.

```{r}
conds <- expand.grid(
  iters = c(250, 1000, 4000), 
  scale = c(0, 0.25, 0.5, 0.75, 1),
  rep = 1:10
)
res <- vector("list", nrow(conds))
chains = 4
for (i in 1:nrow(conds)) {
  iters <- conds[i, "iters"]
  scale <- conds[i, "scale"]
  rep <- conds[i, "rep"]
  r <- array(rnorm(iters * chains), c(iters, chains))
  r[, 1] <- r[, 1] * scale
  rs <- monitor_simple(r)
  res[[i]] <- cbind(iters, scale, rep, rs)
}
res <- bind_rows(res)
```

```{r}
ggplot(data=res, aes(y=zsRhat, x=scale)) + 
  geom_point() + 
  geom_jitter() + 
  facet_grid(. ~ iters) + 
  geom_hline(yintercept = 1.005, linetype = 'dashed') + 
  geom_hline(yintercept = 1) + 
  ggtitle('Split-Rhat')
```

**Result:** Split-Rhat is not able to detect scale differences between chains.

```{r}
ggplot(data=res, aes(y=fzsRhat, x=scale)) + 
  geom_point() + 
  geom_jitter() + 
  facet_grid(. ~ iters) + 
  geom_hline(yintercept = 1.005, linetype = 'dashed') + 
  geom_hline(yintercept = 1) + 
  ggtitle('Folded-split-Rhat')
```

**Result:** Folded-Split-Rhat focuses explicitely on scales 
and can thus detect scale differences.

**Result:** If we use a threshold of $1.01$, we can detect a chain with scale
  less than $3/4$ of the standard deviation of the others. If we use threshold 
  of $1.1$, we can detect a chain with standard deviation less than $1/4$ of 
  the standard deviation of the others.

```{r}
ggplot(data=res, aes(y=zsreff, x=scale)) + 
  geom_point() + 
  geom_jitter() + 
  facet_grid(. ~ iters) + 
  geom_hline(yintercept = c(0,1)) + 
  geom_hline(yintercept = 0.1, linetype = 'dashed') + 
  ggtitle('Relative efficiency (zsreff)') + 
  scale_y_continuous(breaks = seq(0,1,by=0.25))
```

**Result:** The relative efficiency does not see a problem as it
focuses on location differences between chains.

```{r}
ggplot(data=res, aes(y=fzsreff, x=scale)) + 
  geom_point() + 
  geom_jitter() + 
  facet_grid(. ~ iters) + 
  geom_hline(yintercept = c(0,1)) + 
  geom_hline(yintercept = 0.1, linetype = 'dashed') + 
  ggtitle('Folded relative efficiency (fzsreff)') + 
  scale_y_continuous(breaks = seq(0,1,by=0.25))
```

**Result:** The relative efficiency of the standard deviation 
does see the problem as it focuses on scale.

Rank plot visualisation for a case with 4 chains and 250 draws per
chain, and with one chain having a standard deviation of 0.75 as opposed
to a standard deviation of 1 for the other chains.

```{r}
iters = 250
chains = 4
scale = 0.75
r <- array(rnorm(iters * chains), c(iters, chains))
r[, 1] <- r[, 1] * scale
colnames(r) <- 1:4
mcmc_hist_r_scale(r)
```

Folded Rhat is $1.06$, but the rank plot clearly shows where the problem is.

## Cauchy: A distribution with infinite mean and variance

The classic Rhat is based on calculating within and between chain
variances. If the marginal distribution of a chain is such that
the variance is not defined (i.e. infinite), the classic Rhat is not 
well justified. In this section, we will use the Cauchy distribution 
as an example of such distribution. Also in cases where
mean and variance are finite, the distribution can be far from
Gaussian. Especially distributions with very long tails cause
instability for variance and autocorrelation estimates. To alleviate
these problems we will use Split-Rhat for rank-normalized draws. 
Ranks are computed by using average for ties, as this is a deterministic 
procedure and constant chains are mapped to constant chains.

The following Cauchy models are from Michael Betancourt's case study
[Fitting The Cauchy Distribution](https://betanalpha.github.io/assets/case_studies/fitting_the_cauchy.html)

### Nominal parameterization of Cauchy

The nominal cauchy model with direct parameterization is as follows.

```{r}
writeLines(readLines("cauchy_nom.stan"))
```

#### Default Stan options

Run the nominal model:
```{r fit_nom, cache=TRUE, comment=NA, results='hide'}
fit_nom <- stan(file='cauchy_nom.stan', seed=7878, refresh = 0)
```

The diagnostics warn about very large number of transitions after
warmup that exceeded the maximum treedepth, which is likely due to
very long tails of the Cauchy distribution. All chains have low 
estimated Bayesian fraction of missing information also indicating
slow mixing.

MCMC trace for the first parameter looks wild with occasional large values
```{r}
samp <- as.array(fit_nom) 
mcmc_trace(samp[, , 1])
```

Let's check Rhat and relative efficiency diagnostics.

```{r}
res <- monitor_simple(samp[, , 1:50])
which_min_eff <- which(res$zsreff == min(res$zsreff))
```

```{r}
plot_rhat <- function(res) {
  p1 <- ggplot(res, aes(x=par, y=sRhat)) + 
    geom_point() + 
    ggtitle('Classic Split-Rhat') + 
    geom_hline(yintercept = 1.005, linetype = 'dashed') + 
    geom_hline(yintercept = 1) + 
    ylim(c(.99 ,1.26))

  p2 <- ggplot(res, aes(x=par, y=zsRhat)) + 
    geom_point() + 
    ggtitle('Rank normalized split-Rhat') + 
    geom_hline(yintercept = 1.005, linetype = 'dashed') + 
    geom_hline(yintercept = 1) + 
    ylim(c(.99,1.26))
  
  p3 <- ggplot(res, aes(x=par, y=fzsRhat)) + 
    geom_point() + 
    ggtitle('Folded rank split-Rhat') + 
    geom_hline(yintercept = 1.005, linetype = 'dashed') + 
    geom_hline(yintercept = 1) + 
    ylim(c(.99,1.26))
  
  grid.arrange(p1, p2, p3, nrow = 1)
}
```

```{r}
plot_rhat(res)
```

For one parameter, Rhats exceed the classic threshold of 1.1. Depending on the 
Rhat estimate, a few others also exceed the threshold 1.01. The rank normalized
split-Rhat have several values over 1.01. Please note that the classic Rhat is not well defined in this example. 

```{r}
plot_reff <- function(res) {
  ymax <- 2
  ybreaks <- seq(0, ymax, by = 0.25)
  ylimits <- c(0, ymax)
  
  p1 <- ggplot(res, aes(x=par, y=reff)) + 
    geom_point() + 
    ggtitle('Classic rel. efficiency (reff)') + 
    geom_hline(yintercept = c(0,1)) +
    geom_hline(yintercept = 0.1, linetype = 'dashed') + 
    scale_y_continuous(breaks = ybreaks, limits = ylimits) 
  
  p2 <- ggplot(res, aes(x=par, y=zsreff)) + 
    geom_point() + 
    ggtitle('New rel. efficiency (zsreff)') + 
    geom_hline(yintercept = c(0,1)) + 
    geom_hline(yintercept = 0.1, linetype = 'dashed') + 
    scale_y_continuous(breaks = ybreaks, limits = ylimits)
  
  p3 <- ggplot(res, aes(x=par, y=fzsreff)) + 
    geom_point() + 
    ggtitle('Folded rel. efficiency (fzsreff)') + 
    geom_hline(yintercept = c(0,1)) + 
    geom_hline(yintercept = 0.1, linetype = 'dashed') + 
    scale_y_continuous(breaks = ybreaks, limits = ylimits)
  
  p4 <- ggplot(res, aes(x=par, y=medsreff)) + 
    geom_point() + 
    ggtitle('Rel.eff. of median (medsreff)') + 
    geom_hline(yintercept = c(0,1)) + 
    geom_hline(yintercept = 0.1, linetype = 'dashed') + 
    scale_y_continuous(breaks = ybreaks, limits = ylimits)
  
  p5 <- ggplot(res, aes(x=1:50, y=madsreff)) + 
    geom_point() + 
    ggtitle('Rel.eff. of mad (madsreff)') + 
    geom_hline(yintercept = c(0,1)) + 
    geom_hline(yintercept = 0.1, linetype = 'dashed') + 
    scale_y_continuous(breaks = ybreaks, limits = ylimits)
  
  blank <- grid::grid.rect(gp = grid::gpar(col="white"), draw = FALSE)
  grid.arrange(p1, p2, p3, blank, p4, p5, nrow = 2)
}
```

```{r, warning=FALSE}
plot_reff(res) 
```

Both classic and rank normalized relative efficiency estimates have
several near zero values, and the overall samples shouldn't be trusted.

**Result:** Relative efficiency is more sensitive than (rank-normalized) 
split-Rhat to detect problems of slow mixing.

We also check the indicator function `I` and `lp__` and find out that
the relative efficiency for `lp__` is worryingly low.

```{r}
res <- monitor_simple(samp[, , 51:52]) 
cat('I: r_eff = ', round(res['I','zsreff'], 2), '\n')
cat('lp__: r_eff = ', round(res['lp__','zsreff'], 2), '\n')
```

Next, we examine the relative efficiency in different parts of the
posterior by computing the relative efficiency for 
$p(\hat{Q}_\alpha \le x < \hat{Q}_{\alpha+0.05})$, where $\hat{Q}_\alpha$ 
is the empirical $\alpha$-quantile and $\alpha \in (0,0.05,\ldots,0.95)$ 
varies in steps of $5%$.  Each interval contains $5%$ of the draws
and the efficiency measures autocorrelation of an indicator function 
which indicates when the values are inside the specific interval. 
This gives us a local efficiency measure which is does not depend on 
the shape of the distribution.

```{r plot_local_reff}
plot_local_reff <- function(samp = NULL, par = NULL, nalpha = NULL) {
  delta <- 1 / nalpha
  alphas <- seq(0, 1 - delta, by = delta)
  zsreffs <- rep(NA, length(alphas))
  for (i in seq_along(alphas)) {
    alpha <- alphas[i]
    tmp <- samp[, , par]
    I <- tmp >= quantile(tmp, alpha) & tmp < quantile(tmp, alpha + delta)
    rs <- monitor_simple(I)
    zsreffs[i] <- rs$zsreff[1]
  }
  df <- data.frame(
    quantile = seq(0, 1, by = delta), 
    zsreff = c(zsreffs, zsreffs[nalpha])
  )
  ggplot(data=df, aes(x=quantile, y=zsreff)) + 
    geom_step() + 
    geom_hline(yintercept = c(0,1)) + 
    geom_hline(yintercept = 0.1, linetype = 'dashed') + 
    scale_x_continuous(breaks = seq(0,1, by=0.1)) + 
    scale_y_continuous(
      breaks = seq(0, 1, by=0.25), 
      limits = c(0, 1.1)
    )
}
```

```{r}
plot_local_reff(samp = samp, par = 1, nalpha = 20)
```

We see that MCMC is less efficient in the tails.

Alternative way to examine the relative efficiency in different parts
of the posterior, is to compute relative efficiency for 
$p(x < \hat{Q}_{\alpha})$, where $\hat{Q}_\alpha$ is the empirical
$\alpha$-quantile and $\alpha \in (0.025,\ldots,0.975)$. This is
related to the efficiency of quantile estimates. Each interval has
specified proportion of draws, and the efficiency measures
autocorrelation of an indicator function which indicates when the
values are inside the specific interval. This gives us a local
efficiency measure which is indepedent on the shape of the
distribution.

```{r plot_quantile_reff}
plot_quantile_reff <- function(samp = NULL, par = NULL, nalpha=NULL) {
  delta <- 1 / nalpha
  alphas <- seq(delta, 1 - delta, by = delta)
  zsreffs <- rep(NA, length(alphas))
  for (i in seq_along(alphas)) {
    alpha <- alphas[i]
    tmp <- samp[, , par]
    I <- tmp < quantile(tmp, alpha)
    rs <- monitor_simple(I)
    zsreffs[i] <- rs$zsreff[1]
  }
  df <- data.frame(
    quantile = seq(delta, 1 - delta, by = delta), 
    zsreff = zsreffs
  )
  ymax <- max(1, round(max(zsreffs, na.rm = TRUE) + 0.15, 1))
  ggplot(df, aes(x=quantile, y=zsreff)) + 
    geom_point() + 
    geom_hline(yintercept = c(0,1)) + 
    geom_hline(yintercept = 0.1, linetype = 'dashed') + 
    scale_x_continuous(breaks = seq(0,1, by=0.1)) + 
    scale_y_continuous(
      breaks = seq(0, ymax, by = 0.25), 
      limits = c(0, ymax)
    )
}
```

```{r}
plot_quantile_reff(samp = samp, par = 1, nalpha = 40)
```

Rank plot visualisation of x[1], which has relative efficiency 
`r round(res$zsreff[1], 2)`.

```{r}
mcmc_hist_r_scale(samp[, , 1])
```

The chains clearly have different rank plots.

The rank plot visualisation of x[`r which_min_eff`], which has a 
relative efficiency of `r round(min(res$zsreff), 2)`, 
also indicates clear problems with mixing.

```{r}
xmin <- paste0("x[", which_min_eff, "]")
mcmc_hist_r_scale(samp[, , xmin])
```

#### Default Stan options + `max_treedepth=20`

We can try to improve the performance by increasing `max_treedepth`.

```{r fit_nom_td20, cache=TRUE, comment=NA, results='hide'}
fit_nom_td20 <- stan(
  file='cauchy_nom.stan', seed=7878, 
  refresh = 0, control = list(max_treedepth=20)
)
```

MCMC trace for the first parameter looks wild with occasional large values.
```{r}
samp <- as.array(fit_nom_td20)
mcmc_trace(samp[, , 1])
```

```{r}
res <- monitor_simple(samp[, , 1:50])
which_min_eff <- which(res$zsreff == min(res$zsreff))
```

Treedepth exceedence and Bayesian Fraction of Missing Information are
HMC/NUTS specific diagnostics. Let's see what generic Split-Rhat and
relative efficiency can detect. 

We check the diagnostics for all x's.

```{r}
plot_rhat(res)
```

All Rhats are below $1.1$, but many are over $1.01$. Classic Rhat has more
variation than the rank normalized (and the former is not well
defined). The folded rank normalized Rhat shows that there is still
more variation in scale than location between different chains.

```{r, warning=FALSE}
plot_reff(res) 
```

Some classic relative efficiencies (reffs) are close to zero. If we
wouldn't realize that the variance is infinite, we might try to run
longer chains, but in case of infinite variance, zero efficiency is
the truth and longer chains won't help. However other quantities can
be well defined, and that's why it is useful also to look at the
rank normalized version as a generic transformation with finite mean
and variance. The smallest rank-normalized r_eff are around $0.25$,
which is not that bad. There is still large variation between
parameters although they all have the same distribution.

**Result:** Rank normalized relative efficiency is more stable than
classic relative efficiency, which is not well defined for Cauchy.

We check the indicator function `I` and `lp__`. Although increasing 
`max_treedepth` improved diagnostics for `x`, the efficiency for 
`I` and `lp__` didn't change.

```{r}
res <- monitor_simple(samp[, , 51:52])
cat('I: r_eff = ', round(res['I','zsreff'],2), '\n')
cat('lp__: r_eff = ', round(res['lp__','zsreff'],2), '\n')
```

We examine the relative efficiency in different parts of the
posterior, by computing the relative efficiency for 
$p(\hat{Q}_\alpha \le x < \hat{Q}_{\alpha+0.05})$, 
where $\hat{Q}_\alpha$ is $\alpha$-quantile and 
$\alpha \in (0,0.05,\ldots,0.95)$.

```{r}
plot_local_reff(samp = samp, par = 1, nalpha = 20)
```

Increasing `max_treedepth` improved the efficiency slightly in the tails.

We examine the relative efficiency in different parts of the
posterior, by computing the relative efficiency for 
$p(x < \hat{Q}_{\alpha})$, where $\hat{Q}_\alpha$ is $\alpha$-quantile
and $\alpha \in (0,0.025,\ldots,0.975)$ in steps of $0.025$.

```{r}
plot_quantile_reff(samp = samp, par = 1, nalpha = 40)
```

Rank plot visualisation of x[`r which_min_eff`], which has the smallest relative
efficiency `r round(min(res$zsreff))` among the x's. The rank plot indicates clear convergence problems.

```{r}
xmin <- paste0("x[", which_min_eff, "]")
mcmc_hist_r_scale(samp[, , xmin])
```

The rank plot visualisation of `lp__`, which has relative efficiency `round(monitor_simple(samp[,, "lp__"])$zsreff)`, doesn't look so good either.

```{r}
mcmc_hist_r_scale(samp[, , "lp__"])
```

#### Default Stan options + `max_treedepth = 20` + 8 times longer sampling

Let's try running longer chains.
```{r fit_nom_td20l, cache=TRUE, comment=NA, results='hide'}
fit_nom_td20l <- stan(
  file='cauchy_nom.stan', seed=7878, 
  refresh = 0, control = list(max_treedepth=20), 
  warmup=1000, iter = 9000
)
```

MCMC trace for the first parameter looks wild with occasional large values
```{r}
samp <- as.array(fit_nom_td20l)
mcmc_trace(samp[, , 1])
```

```{r}
res <- monitor_simple(samp[, , 1:50])
which_min_eff <- which(res$zsreff == min(res$zsreff))
```

Treedepth exceedence and Bayesian Fraction of Missing Information are
HMC/NUTS specific diagnostics. Let's see what generic Split-Rhat and
relative efficiency can detect. Check the diagnostics for all x's.

```{r}
plot_rhat(res)
```

All Rhats are below $1.01$. Classic Rhat has more variation than the
rank normalized Rhat (but the former is not well defined).

```{r, warning=FALSE}
plot_reff(res) 
```

Most classic relative efficiencies (reffs) are close to zero. If we
wouldn't realize that the variance is infinite, we might try to run
longer chains, but in case of infinite variance zero is the true value
and longer chains don't help. However other quantities can be well
defined, and that's why it can be useful also to look at the rank
normalized version as a generic transformation with finite mean and
variance. The rank-normalized r_eff's start to concentrate around 0.6,
which is quite good. There is still some variation between parameters.

**Result:** Rank normalized relative efficiency is more stable than
classic relative efficiency, which is not well defined for Cauchy.

We check the indicator function `I` and `lp__`:
```{r}
res <- monitor_simple(samp[, , 51:52])
cat('I: r_eff = ', round(res['I','reff'],2), '\n')
cat('lp__: r_eff = ', round(res['lp__','zsreff'],2), '\n')
```

We examine the relative efficiency in different parts of the
posterior, by computing the relative efficiency for 
$p(\hat{Q}_\alpha \le x < \hat{Q}_{\alpha+0.05})$, where 
$\hat{Q}_\alpha$ is $\alpha$-quantile and 
$\alpha \in (0,0.05,\ldots,0.95)$ in steps of $0.05$.

```{r}
plot_local_reff(samp = samp, par = 1, nalpha = 20)
```

Increasing chain length did not seem to change the relative
efficiency. With more draws from the longer chains we can use
finer resolution for the local efficiency estimates.

```{r}
plot_local_reff(samp = samp, par = 1, nalpha = 100)
```

The efficiency far in the tails is worryingly low. We see similar problems 
in the alternative local relative efficiency.

```{r}
plot_quantile_reff(samp = samp, par = 1, nalpha = 100)
```

The rank plot visualisation of x[`r which_min_eff`], which has the smallest relative efficiency `r round(min(res$zsreff), 2)` among the x's.
```{r}
xmin <- paste0("x[", which_min_eff, "]")
mcmc_hist_r_scale(samp[, , xmin])
```

Still problems at the edges.

The rank plot visualisation of `lp__`, which has relative efficiency 
`r round(res["lp__", "zsreff"], 2)`, looks as follows:
```{r}
mcmc_hist_r_scale(samp[, , "lp__"])
```

There seems to be slight problems also in mixing for energy.

<!--
From the above plot, I would naively don't see any problems in mixing.
Or am I just misinterpreting the plot?
-->

### First alternative parameterization of Cauchy

Next we examine alternative parameterization and consider Cauchy as a
scale mixture of Gaussian distributions. The model has two parameters
and the Cauchy distributed x's can be computed from those. In addition
of improved sampling performance, the example illustrates that focus
of diagnostics matter.

```{r}
writeLines(readLines("cauchy_alt_1.stan"))
```

Run the alternative model:

```{r fit_alt1, cache=TRUE, comment=NA, results='hide'}
fit_alt1 <- stan(file='cauchy_alt_1.stan', seed=7878, refresh = 0)
```

```{r}
samp <- as.array(fit_alt1)
res <- monitor_simple(samp[, , 101:150])
```

There are no warnings, and the sampling is much faster.

```{r}
plot_rhat(res)
```

All Rhats are below 1.01. Classic Rhat's look also good even if the
classic is not well defined for Cauchy distribution.

```{r, warning=FALSE}
plot_reff(res) 
```

**Result:** Rank normalized r_eff's have less variation than classic 
one which is not well defined for Cauchy.

We check the indicator function `I` and `lp__`:
```{r}
res <- monitor_simple(samp[, , 151:152])
cat('I: r_eff = ', round(res['I','zsreff'], 2), '\n')
cat('lp__: r_eff = ', round(res['lp__','zsreff'], 2), '\n')
```

The relative efficiency for these are also much better than with
nominal parameterization.

We examine the relative efficiency in different parts of the
posterior, by computing the relative efficiency for 
$p(\hat{Q}_\alpha \le x < \hat{Q}_{\alpha + 0.05})$, where 
$\hat{Q}_\alpha$ is $\alpha$-quantile and $\alpha \in (0,0.05,\ldots,0.95)$.

```{r}
samp <- as.array(fit_alt1)
plot_local_reff(samp = samp, par = 101, nalpha = 20)
```

The relative efficiency is good in all parts of the posterior.

We examine the relative efficiency in different parts of the
posterior, by computing the relative efficiency for 
$p(x < \hat{Q}_{\alpha})$, where $\hat{Q}_\alpha$ is $\alpha$-quantile
and $\alpha \in (0,0.025,\ldots,0.975)$.

```{r}
samp <- as.array(fit_alt1)
plot_quantile_reff(samp = samp, par = 101, nalpha = 40)
```

```{r}
res <- monitor_simple(samp[, , 101:150])
which_min_eff <- which(res$zsreff == min(res$zsreff))
res1 <- monitor_simple(samp[, , 1:50])
res2 <- monitor_simple(samp[, , 51:100])
```

```{r}
cat('Mean r_eff for xs = ' , round(mean(res[,'zsreff']), 2), '\n')
cat('Mean fr_eff for xs = ' , round(mean(res[,'fzsreff']), 2), '\n')
cat('Mean r_eff for x_as = ' , round(mean(res1[,'zsreff']), 2), '\n')
cat('Mean r_eff for x_bs = ' , round(mean(res2[,'zsreff']), 2), '\n')
```

**Result:** We see that the relative efficiency of the interesting $x$
can be different from the relative efficiency of the parameters that
used to compute it. Also r_eff we are computing is for rank
normalized and to compute Monte Carlo errors, the final target
function should be taken into account, but we just want to avoid the
complication of the potentially non-existing moments at this point,
and come back later in other paper how to use Pareto diagnostic to
check whether necessary moments exist to compute the desired
expectations and their Monte Carlo errors.

<!--
The above result sections is hard to read and understand.
Can we break this down a little bit?
-->

Rank plot visualisation of x[`r which_min_eff`], which has the smallest relative
efficiency `r round(min(res$zsreff), 2)` among x's.

```{r}
xmin <- paste0("x[", which_min_eff, "]")
mcmc_hist_r_scale(samp[, , xmin])
```

Looks better than for the nominal parameterization.

Rank plot visualisation of `lp__`, which has a relative efficiency of `r round(monitor_simple(samp[,, "lp__"]), 2)`.
```{r}
mcmc_hist_r_scale(samp[, , "lp__"])
```

Looks better than for the nominal parameterization. It would be useful
to have the reference plotted.

### Another alternative parameterization of the Cauchy distribution

Another alternative parameterization is univariate transformation as
shown in the following code (3rd alternative in Michael Betancourt's case study).

```{r}
writeLines(readLines("cauchy_alt_3.stan"))
```

Run the alternative model:
```{r fit_alt2, cache=TRUE, comment=NA, results='hide'}
fit_alt3 <- stan(file='cauchy_alt_3.stan', seed=7878, refresh = 0)
```

There are no warnings, and the sampling is much faster than for the
nominal model.

```{r}
samp <- as.array(fit_alt3)
res <- monitor_simple(samp[, , 51:100])
```

```{r}
plot_rhat(res)
```

All Rhats except some folded Rhats are below 1.01. Classic Rhat's look
also good even if it is not well defined for the Cauchy distribution.

```{r, warning=FALSE}
plot_reff(res) 
```

**Result:** Rank normalized relative efficiencies have less variation
than classic ones which is not well defined for Cauchy. 
Rank normalized relative efficiencies are slightly larger than 1, 
which is possible for antithetic Markov chains which have negative 
correlation for odd lags.

We check the indicator function `I` and `lp__`:
```{r}
res <- monitor_simple(samp[, , 101:102])
cat('I: r_eff = ', round(res['I','zsreff'],2), '\n')
cat('lp__: r_eff = ', round(res['lp__','zsreff'],2), '\n')
```

The relative efficiency for these are also much better than with
the nominal parameterization.

We examine the relative efficiency in different parts of the
posterior, by computing the relative efficiency for 
$p(\hat{Q}_\alpha \leq x < \hat{Q}_{\alpha+0.05})$, where 
$\hat{Q}_\alpha$ is $\alpha$-quantile and 
$\alpha \in (0,0.05,\ldots,0.95)$.

```{r}
samp <- as.array(fit_alt3)
plot_local_reff(samp = samp, par = 51, nalpha = 20)
```

We examine the relative efficiency in different parts of the
posterior, by computing the relative efficiency for 
$p(x < \hat{Q}_{\alpha})$, where $\hat{Q}_\alpha$ is 
$\alpha$-quantile and $\alpha \in (0,0.025,\ldots,0.975)$.

```{r}
samp <- as.array(fit_alt3)
plot_quantile_reff(samp = samp, par = 51, nalpha = 40)
```

The relative efficiency in tails is worse than for the first
alternative parameterization, although it's still better than for the
nominal model.

```{r}
res <- monitor_simple(samp[, , 51:100])
which_min_eff <- which(res$zsreff == min(res$zsreff))
res1 <- monitor_simple(samp[, , 1:50])
cat('Mean r_eff for xs = ' , round(mean(res[,'zsreff']),2), '\n')
cat('Mean fr_eff for xs = ' , round(mean(res[,'fzsreff']),2), '\n')
cat('Mean r_eff for x_as = ' , round(mean(res1[,'zsreff']),2), '\n')
```

**Result:** When the transformation is univariate and bijective, the
rank normalized Rhat and r_eff are transformation invariant.

Naturally when computing the desired expectations, the final target
function should be taken into account, but we just want to avoid the
complication of the potentially non-existing moments at this point,
and come back later in other paper how to use Pareto diagnostic to
check whether necessary moments exist to compute the desired
expectations and their Monte Carlo errors.

Rank plot visualisation of x[`r which_min_eff`], which has the smallest relative
efficiency of `r round( min(res$zsreff), 2)` among x's.

```{r}
xmin <- paste0("x[", which_min_eff, "]")
mcmc_hist_r_scale(samp[, , xmin])
```

Nothing special.

Rank plot visualisation of `lp__`, which has relative efficiency 0.33.
```{r}
mcmc_hist_r_scale(samp[, , "lp__"])
```

Nothing special.

### Half-Cauchy with nominal parameterization

Half-Cauchy priors are common and, for example, in Stan usually set
using the nominal parameterization. However, when the constraint 
`<lower=0>` is used, Stan does the sampling automatically
in the unconstrained `log(x)` space, which changes the geometry
crucially.

```{r}
writeLines(readLines("half_cauchy_nom.stan"))
```

Run the half-Cauchy with nominal parameterization (and positive constraint).

```{r fit_half_nom, cache=TRUE, comment=NA, results='hide'}
fit_half_nom <- stan(file='half_cauchy_nom.stan', seed=7878, refresh = 0)
```

There are no warnings, and the sampling is much faster than for the
Cauchy nominal model.

```{r}
samp <- as.array(fit_half_nom)
res <- monitor_simple(samp[, , 1:50])
```

```{r}
plot_rhat(res) 
```

All Rhats are below 1.01. Classic Rhat's look also good even if the
classic is not well defined for half-Cauchy distribution.

```{r, warning=FALSE}
plot_reff(res)  
```

**Result:** Rank normalized r_eff's have less variation than classic
one which is not well defined for Cauchy. Rank normalized r_eff's
are much larger than 1, which is possible for antithetic Markov
chains which have negative correlation for odd lags.

Due to constraint `<lower=0>`, the sampling was made in `log(x)` space,
and we can also check the performance in that space.
```{r}
res <- monitor_simple(samp[, , 1:50])
```

```{r, warning=FALSE}
plot_reff(res) 
```

$log(x)$ is quite close to Gaussian, and thus classic r_eff is also
close to rank normalized r_eff which is exactly same as for $x$ as the
rank normalized version is invariant to bijective transformations.

**Result:** Rank normalized relative efficiency is close to classic 
relative efficiency for transformations which make the distribution 
close to Gaussian.

We check the indicator function `I` and `lp__`:
```{r}
res <- monitor_simple(samp[, , 51:52])
which_min_eff <- which(res$zsreff == min(res$zsreff))
cat('I: r_eff = ', round(res['I','zsreff'],2), '\n')
cat('lp__: r_eff = ', round(res['lp__','zsreff'],2), '\n')
```

**Result:** Relative efficiency for $p(x_1 < 1)$ is larger than 1. 
This is possible for antithetic Markov chains, which have negative
correlation for odd lags.

We examine the relative efficiency in different parts of the
posterior, by computing the relative efficiency for $p(\hat{Q}_\alpha
\le x < \hat{Q}_{\alpha+0.05})$, where $\hat{Q}_\alpha$ is
$\alpha$-quantile and $\alpha \in (0,0.05,\ldots,0.95)$.

```{r}
samp <- as.array(fit_half_nom)
plot_local_reff(samp = samp, par = 1, nalpha = 20)
```

The relative efficiency is good overall, with only a small dip in tails.

We examine the relative efficiency in different parts of the
posterior, by computing the relative efficiency for 
$p(x < \hat{Q}_{\alpha+0.05})$, where $\hat{Q}_\alpha$ is 
$\alpha$-quantile and $\alpha \in (0,0.025,\ldots,0.975)$.

```{r}
samp <- as.array(fit_half_nom)
plot_quantile_reff(samp = samp, par = 1, nalpha = 40)
```

Rank plot visualisation of x[`r which_min_eff`], which has the smallest relative
efficiency 1.0 among x's.
```{r}
xmin <- paste0("x[", which_min_eff, "]")
mcmc_hist_r_scale(samp[, , xmin])
```

Looks like there are differences in scale of chains.

Rank plot visualisation of `lp__`, which has relative efficiency 0.33.
```{r}
mcmc_hist_r_scale(samp[, , "lp__"])
```

Maybe small differences in scales, but it's difficult to know whether
small variation from uniform is relevant.

### Alternative parameterization of half-Cauchy

```{r}
writeLines(readLines("half_cauchy_alt.stan"))
```

Run half-Cauchy with alternative parameterization
```{r fit_half_reparam, cache=TRUE, comment=NA, results='hide'}
fit_half_reparam <- stan(file = 'half_cauchy_alt.stan', seed=7878, refresh = 0)
```

There are no warnings, and the sampling is as fast for the half-Cauchy
nominal model.

```{r}
samp <- as.array(fit_half_reparam)
res <- monitor_simple(samp[, , 101:150])
```

```{r}
plot_rhat(res)
```

All Rhats are below 1.01. Classic Rhat's look also good even if the
classic is not well defined for half-Cauchy distribution.

```{r, warning=FALSE}
plot_reff(res) 
```

**Result:** Rank normalized relative efficiencies have less variation 
than classic ones which is not well defined for Cauchy. 
Based on rank normalized relative efficiencies the alternative 
parameterization has much lower efficiency than the nominal 
parameterization with constraint `<lower=0>`.

We check the indicator function `I` and `lp__`:
```{r}
res <- monitor_simple(samp[, , 151:152])
cat('I: r_eff = ', round(res['I','zsreff'],2), '\n')
cat('lp__: r_eff = ', round(res['lp__','zsreff'],2), '\n')
```

We examine the relative efficiency in different parts of the
posterior, by computing the relative efficiency for $p(\hat{Q}_\alpha
\le x < \hat{Q}_{\alpha+0.05})$, where $\hat{Q}_\alpha$ is
$\alpha$-quantile and $\alpha \in (0,0.05,\ldots,0.95)$.

```{r}
samp <- as.array(fit_half_reparam)
plot_local_reff(samp = samp, par = 101, nalpha = 20)
```

We examine the relative efficiency in different parts of the
posterior, by computing the relative efficiency for $p(x <
\hat{Q}_{\alpha+0.05})$, where $\hat{Q}_\alpha$ is $\alpha$-quantile
and $\alpha \in (0,0.025,\ldots,0.975)$.

```{r}
samp <- as.array(fit_half_reparam)
plot_quantile_reff(samp = samp, par = 1, nalpha = 40)
```

The relative efficiency near zero is much worse than for the
half-Cauchy with nominal parameterization and constraint `<lower=0>`.

Rank plot visualisation of `x[15]`, which has the smallest relative
efficiency 0.30 among `x`'s.
```{r}
mcmc_hist_r_scale(samp[, , "x[15]"])
```

The rank plot is different from uniform, which is expected with lower
relative efficiency, but would require a reference.

Rank plot visualisation of `lp__`, which has relative efficiency 0.24.
```{r}
mcmc_hist_r_scale(samp[, , "lp__"])
```

The rank plot is different from uniform, which is expected with lower
relative efficiency, but would require a reference.

## Hierarchical model: Eight Schools

The models are from Michael Betancourt's case study on
[Diagnosing Biased Inference with Divergences](http://mc-stan.org/users/documentation/case-studies/divergences_and_bias.html).

### A Centered Eight Schools model

```{r, comment=NA}
writeLines(readLines("eight_schools_cp.stan"))
```

#### Centered parameterization default MCMC options

Run centered parameterization model with default MCMC options.
```{r fit_cp, cache=TRUE, comment=NA, results='hide'}
input_data <- read_rdump("eight_schools.data.R")
fit_cp <- stan(
  file='eight_schools_cp.stan', data=input_data,
  iter=2000, chains=4, seed=483892929, refresh=0
)
```

We do observe divergent transitions, which is sensitive diagnostic for
this model and we observer also low BFMI, but let's assume we were
investigating a MCMC algorithm without such diagnostic.

```{r}
sel <- c(
  'sRhat', 'zsRhat', 'fzsRhat', 'reff',
  'zsreff', 'fzsreff', 'medsreff', 'madsreff'
)
res <- monitor_simple(fit_cp)
round(res[, sel], 2)
```

Some rank-normalized split-Rhats are larger than 1.01. Relative
efficiencies for tau and lp__ are less than 10%, which is worry-some
and longer chains should be run.

We examine the relative efficiency of probabilities between quantiles
(0, 0.05, ..., 1) for `tau`.

```{r}
samp <- as.array(fit_cp)
plot_local_reff(samp = samp, par = "tau", nalpha = 20)
```

We see that MCMC has difficulties in exploring small tau values. As
the efficiency for estimating small tau values is practically zero, we
may assume that we may miss substantial amount of posterior mass and
get biased estimates.

We examine the relative efficiency of cumulative probabilities below
quantiles $(0.025, ..., 0.975)$ for `tau`.

```{r}
samp <- as.array(fit_cp)
plot_quantile_reff(samp = samp, par = 2, nalpha = 40)
```

Rank plot visualisation of `tau`, which has the smallest relative
efficiency 0.05 among parameters.

```{r}
mcmc_hist_r_scale(samp[, , "tau"])
```

We observe clear problems.

Rank plot visualisation of `lp__`, which has relative efficiency 0.05.
```{r}
mcmc_hist_r_scale(samp[, , "lp__"])
```

We observe clear problems.

#### Centered parameterization with longer chains

Low r_eff can be sometimes compensated with longer chains. Let's check
10 times longer chain.

```{r fit_cp2, cache=TRUE, comment=NA, results='hide'}
fit_cp2 <- stan(
  file='eight_schools_cp.stan', data=input_data,
  iter=20000, chains=4, seed=483892929, refresh=0
)
```

```{r}
res <- monitor_simple(fit_cp2)
round(res[, sel], 2)
```

Some rank-normalized split-Rhats are still larger than 1.01. Relative
efficiencies for tau and `lp__` are around 1%. A drop in the relative
efficiency when increasing the number of iterations indicates serious
problems in mixing.

We examine the relative efficiency of probabilities between quantiles
$(0, 0.05, ..., 1)$ for `tau`.

```{r}
samp <- as.array(fit_cp2)
plot_local_reff(samp = samp, par = 2, nalpha = 50)
```

We see that MCMC has difficulties in exploring small `tau` values. As
the efficiency for estimating small `tau` values is practically zero, 
we may assume that we may miss substantial amount of posterior mass and
get biased estimates.

We examine the relative efficiency of cumulative probabilities below
quantiles $(0.01, ..., 0.99)$ for `tau`.

```{r}
samp <- as.array(fit_cp2)
plot_quantile_reff(samp = samp, par = 2, nalpha = 100)
```

We examine also other parameters, like `mu` in the following plot.
```{r}
samp <- as.array(fit_cp2)
plot_local_reff(samp = samp, par = 1, nalpha = 50)
```

There are gaps of poor efficiency which indicates sticking of MCMC,
but the sticking doesn't occur for any specific range of values of `mu`.

We examine also other parameters, like `mu` in the following plot.
```{r}
samp <- as.array(fit_cp2)
plot_quantile_reff(samp = samp, par = 1, nalpha = 100)
```

Rank plot visualisation of `tau`, which has the smallest relative
efficiency 0.01 among parameters.
```{r}
mcmc_hist_r_scale(samp[, , "tau"])
```

We observe clear problems to sample small values of tau.

Rank plot visualisation of lp__, which has relative efficiency 0.01.
```{r}
mcmc_hist_r_scale(samp[, , "lp__"])
```

We observe clear problems sampling different energy levels.

#### Centered parameterization very long chains

And for further evidence let's check 100 times longer chains than the default.
```{r fit_cp3, cache=TRUE, comment=NA, results='hide'}
fit_cp3 <- stan(
  file='eight_schools_cp.stan', data=input_data,
  iter=200000, chains=4, seed=483892929, refresh=0
)
```

```{r}
res <- monitor_simple(fit_cp3)
round(res[, sel], 2)
```

Some rank-normalized split-Rhats are still larger than 1.01. Relative
efficiencies for tau and lp__ are around 0% and relative efficiencies
for other parameters are also getting smaller. Drop in the relative
efficiency when the the sample size was increased indicates serious
problems in mixing.

We examine the relative efficiency of probabilities between quantiles
$(0, 0.05, ..., 1)$ for tau.
```{r}
samp <- as.array(fit_cp3)
plot_local_reff(samp = samp, par = 2, nalpha = 100)
```

We see that MCMC has difficulties in exploring small tau values. As
the efficiency for estimating small tau values is practically zero, we
may assume that we may miss substantial amount of posterior mass and
get biased estimates. It is good to note that the low efficiency for
small tau values is due to too large step size, and the low efficiency
for large tau values is partially due to too small step size, but even
more importantly due to the random walk in energy space.

We examine the relative efficiency of cumulative probabilities below
quantiles (0.01, ..., 0.99) for tau.
```{r}
samp <- as.array(fit_cp3)
plot_quantile_reff(samp = samp, par = 2, nalpha = 100)
```

Rank plot visualisation of tau, which has the smallest relative
efficiency 0.00 among parameters.
```{r}
mcmc_hist_r_scale(samp[, , 2])
```

We observe clear problems to sample small values of tau and even with 100000
draws per chain, the plots don't get crowded as traceplots would.

Rank plot visualisation of lp__, which has relative
efficiency 0.01.
```{r}
mcmc_hist(r_scale(samp[,,11]), breaks = seq(0,prod(dim(samp)[1:2]),by=prod(dim(samp)[1:2])/100)+0.5)
```

We observe clear problems sampling different energy levels, and
the low efficiency for lp__ comes just from slow random walk
(autocorrelation for chains 1 and 2 goes to zero after 1000 lags, and
for chains 3 and 4 after 5000 lags).


### Non-centered Eight Schools model

For hierarchical models, non-centered parameterization often works better
```{r, comment=NA}
writeLines(readLines("eight_schools_ncp.stan"))
```

#### Non-centered parameterization default MCMC options

Run non-centered parameterization model with default options.
```{r fit_ncp, cache=TRUE, comment=NA, results='hide'}
input_data <- read_rdump("eight_schools.data.R")
fit_ncp <- stan(
  file='eight_schools_ncp.stan', data=input_data,
  iter=2000, chains=4, seed=483892929, refresh=0
)
```

We still observe some divergent transitions with the default
adapt_delta and no divergences with adapt_delta=0.95. Let's analyze
both samples.

```{r}
res <- monitor_simple(fit_ncp)
round(res[, sel], 2)
```

All Rhats are close to 1, and relative efficiencies are good.

We examine the relative efficiency of probabilities between quantiles
$(0, 0.05, ..., 1)$ for tau.
```{r}
samp <- as.array(fit_ncp)
plot_local_reff(samp = samp, par = 2, nalpha = 20)
```

Small tau values are still more difficult to explore, but the relative
efficiency is in a good range.

We examine the relative efficiency of probabilities below quantiles
(0.025, ..., 0.975) for tau.
```{r}
samp <- as.array(fit_ncp)
plot_quantile_reff(samp = samp, par = 2, nalpha = 40)
```

Rank plot visualisation of tau, which has the smallest relative
efficiency 0.57 among parameters.

```{r}
mcmc_hist_r_scale(samp[, , 2])
```

We can see problems in sampling small values of tau.

Rank plot visualisation of lp__, which has relative efficiency 0.01.
```{r}
mcmc_hist_r_scale(samp[, , 19])
```

We observe clear problems sampling different energy levels.

#### Non-centered parameterization default MCMC options plus `adapt_delta=0.95`

Next we examine the same model but with higher `adapt_delta=0.95`.
```{r fit_ncp2, cache=TRUE, comment=NA, results='hide'}
fit_ncp2 <- stan(
  file='eight_schools_ncp.stan', data=input_data,
  iter=2000, chains=4, control=list(adapt_delta = 0.95), 
  seed=483892929, refresh=0
)
```

We get zero divergences with `adapt_delta = 0.95`.

```{r}
res <- monitor_simple(fit_ncp2)
round(res[, sel], 2)
```

All Rhats are close to 1, and relative efficiencies are good and
slightly better than with the default `adapt_delta`.

We examine the relative efficiency of probabilities between quantiles
$(0, 0.05, ..., 1)$ for `tau`.
```{r}
samp <- as.array(fit_ncp2)
plot_local_reff(samp = samp, par = 2, nalpha = 20)
```

Small tau values are still more difficult to explore, but the relative
efficiency is in a good range.

Check with a finer resolution
```{r}
samp <- as.array(fit_ncp2)
plot_local_reff(samp = samp, par = 2, nalpha = 50)
```

Also with a finer resolution the efficiency of MCMC is good for
different values of tau.

We examine the relative efficiency of probabilities below quantiles
$(0.025, ..., 0.975)$ for `tau`.
```{r}
samp <- as.array(fit_ncp2)
plot_quantile_reff(samp = samp, par = 2, nalpha = 40)
```

Rank plot visualisation of `tau`, which has the smallest relative
efficiency 0.62 among parameters.
```{r}
mcmc_hist_r_scale(samp[, , 2])
```

Higher adapt_delta seems to have improved sampling of small values.

Rank plot visualisation of `lp__`, which has relative efficiency 0.44.
```{r}
mcmc_hist_r_scale(samp[, , 19])
```

We observe clear problems sampling different energy levels.

#### Non-centered parameterization default MCMC options + adapt_delta=0.95 + longer chains

If in doubt, we can run longer chains.
```{r fit_ncp3, cache=TRUE, comment=NA, results='hide'}
input_data <- read_rdump("eight_schools.data.R")
fit_ncp3 <- stan(
  file='eight_schools_ncp.stan', data=input_data,
  iter=20000, chains=4, control=list(adapt_delta = 0.954), 
  seed=483892929, refresh=0
)
```

```{r}
res <- monitor_simple(fit_ncp3)
round(res[, sel], 2)
```

All Rhats are close to 1, and relative efficiencies are good and
slightly better than with the default adapt_delta. Relative efficiency
is similar as for shorter chain, which means that running longer
chains is giving us higher effective sample size.

We examine the relative efficiency of probabilities between quantiles
$(0, 0.05, ..., 1)$ for `tau`.
```{r}
samp <- as.array(fit_ncp3)
plot_local_reff(samp = samp, par = "tau", nalpha = 100)
```

Small `tau` values are still more difficult to explore, but the relative
efficiency is in a good range.

We examine the relative efficiency of probabilities below quantiles
$(0.01, ..., 0.99)$ for `tau`.
```{r}
samp <- as.array(fit_ncp2)
plot_quantile_reff(samp = samp, par = "tau", nalpha = 100)
```

Rank plot visualisation of `tau`, which has the smallest relative
efficiency of `r round(monitor_simple(samp[, , "tau"])$zsreff, 2)` among parameters.
```{r}
mcmc_hist_r_scale(samp[, , "tau"])
```

Higher adapt_delta seems to have improved sampling of small values.

Rank plot visualisation of `lp__`, which has a relative efficiency of `r round(monitor_simple(samp[, , "lp__"])$zsreff, 2)`.
```{r}
mcmc_hist_r_scale(samp[, , "lp__"])
```

We observe clear problems sampling different energy levels.

## NUTS and effective sample size

We have already seen that the relative efficiency of HMC/NUTS can be
higher than with independent draws. The next example illustrates
interesting relative efficiency phenomena due to properties of the
HMC/NUTS algorithm.

```{r}
writeLines(readLines("normal.stan"))
```

```{r fit_n, cache=TRUE, comment=NA, results='hide'}
fit_n <- stan(
  file='normal.stan', data=data.frame(J=16),
  iter=20000, chains=4, seed=483892929, refresh=0
)
```

```{r}
samp <- as.array(fit_n)
res <- monitor_simple(samp)
round(res[, sel], 2)
```

The relative efficiency for all x is larger than 
`r round(min(res[1:16, "zsreff"]), 2)`. 
The relative efficiency for `lp__` is however only 
`r round(res["lp__", "zsreff"], 2)`. If we take a look at all
the Stan examples in this notebook, we see that the relative efficiency 
for `lp__` is always below 0.5. `lp__` correlates strongly with the total
energy in HMC, and that total energy is sampled using random walk
proposal once per iteration and thus it's likely that `lp__` has some
random walk behavior leading to autocorrelation and relative
efficiency below 1. On the other hand, No-U-Turn behavior and sampling
from the trajectory may make the Markov chain to be antithetic with
negative odd lag correlation and relative efficiency higher than 1
for some of its parameters-

Let's check the local relative efficiency.
```{r}
plot_local_reff(samp = samp, par = 1, nalpha = 100)
```

The relative efficiency for probability estimate for a small interval
is close to 1 with a slight drop in the tails. This is a good result,
but far from the relative efficiency for mean estimate.

Let's check the relative efficiency of cumulative probabilities.
```{r}
plot_quantile_reff(samp = samp, par = 1, nalpha = 100)
```

The total energy of HMC should affect how far in the tails a chain in
one iteration can go. Fat tails of the target have high energy, and
thus only chains with high total energy can reach there. This will
suggest that the random walk in total energy would cause random walk
in variance of x. Let's check the second moment of x.
```{r}
samp <- as.array(fit_n, pars = "x")^2
res <- monitor_simple(samp)
round(res[, sel], 2)
```

Mean of the relative efficiencies for $x_j^2$ is 
`r round(mean(res$zsreff), 2)`, which is quite
close to the relative efficiency for `lp__`. This is not that
surprising as the potential energy in normal model is relative to
$\sum_{j=1}^J x_j^2$.

Let's check the local relative efficiency
```{r}
plot_local_reff(samp = samp, par = 1, nalpha = 100)
```

The relative efficiency is mostly a bit below 1, but for the right
tail of $x_1^2$ the relative efficiency drops. This likely due to
only some iterations have high enough total energy to obtain draws
from the high energy part of the tail.

Let's check the relative efficiency of cumulative probabilities
```{r}
plot_quantile_reff(samp = samp, par = 1, nalpha = 100)
```

We can see the correlation between `lp__` and magnitude of `x[1]` in
the following plot.
```{r}
samp <- as.array(fit_n)
qplot(
  as.vector(samp[, , "lp__"]),
  abs(as.vector(samp[, , "x[1]"]))
) + 
  xlab('lp__') + 
  ylab('x[1]')
```

Low `lp__` corresponds to high energy and more variation in `x[1]`, and
high `lp__` corresponds to low energy and small variation in `x[1]`. Finally $\sum_{j=1}^J x_j^2$ is perfectly correlated with `lp__`.
```{r}
qplot(
  as.vector(samp[, , "lp__"]),
  as.vector(apply(samp[, , 1:16]^2, 1:2, sum))
) + 
  xlab('lp__') + 
  ylab('sum(x^2)')
```

This shows that even if we get high relative efficiency estimates for
parameters, it is important to look at the relative efficiency of
`lp__` as it can indicate problems of sampling in tails, which would
affect the accuracy of variance and tail quantile estimates.

# Original Computing Environment

```{r, comment=NA}
writeLines(readLines(file.path(Sys.getenv("HOME"), ".R/Makevars")))
```

```{r, comment=NA}
devtools::session_info("rstan")
```

# References
